---
title: "How to keep it togetheR"
subtitle: "List columns in a dataframe"
author: "Marc A.T. Teunis, Ph.D."
date: "`r Sys.Date()`"
output: ioslides_presentation
---

<style>

.footer {
    font-size: 16pt;
    color: Black;
    font-family: 'Calibri'
    content: attr(label);
    position: absolute;
    margin-left: auto;
    margin-right: auto;
    left: 0;
    right: 0;
    text-align:center;
    bottom: 10px;
    line-height: 1.9;
    display: block;
}

</style>

<div class="footer" style="margin-top;font-size:120%;"> 
Using list-columns - R Cafe UU, 28 January 2019 </div>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE, 
                      error = FALSE, 
                          fig.width = 6,
                          fig.height = 8)
```

```{r, root, include=FALSE}
require("rprojroot") || utils::install.packages("rprojroot")
library(rprojroot)
root <- find_root_file(criterion = is_rstudio_project)

```

## Prerequisites

 - Intro base R
 - Data wrangling
 - Visualizations
 - Writing functions
 - Loops 
 - Pipes (`%>%`)

## Credits

Much of this material has been derived in one way or the other from the work of Hadley Wickham and Garret Grolemund and many others. For a more elaborate reference list see the REFERENCES.txt file in the project root.

Thanks to Hadley & Garret for writing the book "R for Data Science" http://r4ds.had.co.nz/ and for their work in general to innovate the R world.

Work on integration of Git/Github with R/RStudio is thoroughly and wit-fully documented by Jenny Brian. I also very much appreciate her work on the use of RMarkdown. See also:

https://github.com/jennybc/happy-git-with-r & http://stat545.com/block007_first-use-rmarkdown.html 

## Contents

List columns and managing many models with `{tidyverse}`

 1. Quick R recap
 2. Loops in R
 3. Using list colums (in a dataframe)
 4. Implementing list columns to create models and graphs

## Functional programming in R

 - Writing your own functions
 - Automate routine tasks
 - Prerequisite for writing R packages
 - Prerequisite for increasing reproducibility
 
## Iterations

 - Iterate a function over a number of elements/objects
 - `for`-loops are not bad
 - `map` (or `apply`) family of functions are specials for iteration
 - Iterations and writing functions go hand-in-hand
 
## Packages
```{r, packages}
library(tidyverse)
library(modelr)
library(lubridate)
library(broom)
```

# Case data

## Data origin

Whooping cough outbreaks from The World Health Organization

http://data.euro.who.int/cisid/?TabID=463987

See for variable and data collection also: 

http://ecdc.europa.eu/sites/portal/files/documents/Pertussis%20AER.pdf

## Data gathering
The data used in this presentation has been constructed from the interactive database tool hosted at: http://data.euro.who.int/cisid/ 
A selection was made for all available countries and all avaialable years, for the number of cases reported to the WHO for Whooping Cough - __Bordetella pertussis__ - infections.
The file is available in CSV format in `./data/CISID_pertussis_10082018.csv`

## Data characteristics

- The first few lines of the file have comments indicated with `#`
- There are 53 countries in the dataset
- There is data from 1991 to 2013 
- The data is not in `tidy` format
- Missing values are indicated with _`NA`_

## **Discuss with your neighbour**

 - What makes this data untidy?
 - Why is it not a good idea to have a column name starting with a digit?

## Read data into R
```{r, read_data}
pertussis_data <- read_csv(
  file = file.path(root, 
  "data", 
  "CISID_pertussis_10082018.csv"),
  comment = "#", 
  n_max = 53, 
  na = c("", " ")
  )
```

## Inspect data
```{r, inspect_data}
head(pertussis_data, n = 2)
names(pertussis_data)
```

## The pipe

**%>%** (`infix function from the {magrittr} package`)

`some_object %>% function_that_does_something_to_the_object()`

equivalent to:

`function_that_does_something_to_the_object(some_object)`

Avoid recursive use of brackets, use the pipe

**"Write code for humans and computers"**

## Tidy the data
```{r}
source(file = file.path(root,
                        "code",
                        "do_tidy_pertussis.R"))
```

# Recap 'R Objects'

## R vectors

 - Vectors are the building blocks of everything that is `R`
 - There are two flavours: `atomic` vectors and `recursive` vectors (lists)
 - `atomic` vectors are homogenous, `recursive vectors` can be heterogenous

## Atomic vectors
 
 - The shortest vector is `NULL`, the longest is `inf`, the weirdest is `Nan`
```{r}
(28/0) / (28/0)
```
 - A single element is also a vector
 - Columns of a tabular data structure are also vectors

## Proof of the pudding is in the eating 

Atomic vectors
```{r, atomic_vectors}
null <- NULL
is.vector(null)
inf <- Inf
is.vector(inf)
nan <- NaN
is.vector(nan)
single_element <- 1
is.vector(single_element)
```

## Recursive vectors
```{r, recursive_vectors}
table <- tribble(
  ~ a, ~b, ~c, ~d,
  "x", 1, TRUE, 1.45, 
  "y", 2, FALSE, 3.88,
  "z", 3, TRUE, 33.5
  ) %>%
  as_tibble()
table
```

## Checking characteristics of `table`
```{r}
is.vector(table)
is.vector(table$a)
is.atomic(table)
is.atomic(table$c)
is.recursive(table)
class(table)
typeof(table)
## So dataframes are a special kind of lists?!? 
```

## **Discuss with your neighbour** 
What is going on with the code below?
```{r} 
table[1]
table[[1]]
```

## Data frames, Tibbles, and List Columns
```{r}
data.frame(a = c(1, 2, 3.14), 
           b = c("a", "b", "c"), 
           c = c(TRUE, FALSE, FALSE)) %>%
  print()
```

## A list-column in a 'normal' dataframe
```{r}
data.frame(a = c(1, 2, 3.14), 
           b = c("a", "b", "c"), 
           c = c(TRUE, FALSE, FALSE),
           d = list(p = 1:3, q = TRUE, r = 0L)) %>%
  print()
```

## Add list column with `$`
```{r}
z <- data.frame(a = c(1, 2, 3.14), 
                b = c("a", "b", "c"), 
                c = c(TRUE, FALSE, FALSE))
z$d <- list(p = 1:30, q = TRUE, r = 0L)
print(z)
## prints messy to the Console
```

## `tibble` for nicer printing
```{r}
library(tibble)
z <- as_tibble(z)
print(z)
```

## Create tibble directly
```{r}
tibble(a = c(1, 2, 3.14), 
       b = c("a", "b", "c"), 
       c = c(TRUE, FALSE, FALSE),
       d = list(p = 1:3, q = TRUE, r = 0L))
```

## Use visual `r`epresentation of the datframe for illustration/education
```{r}
tribble( ~a,  ~b,     ~c,      ~d,
         1,   "a",    TRUE,    p = 1:3,
         2,   "b",    FALSE,   q = TRUE,
      3.14,   "c",    FALSE,   r = 0L)
```

## Lists
 - List can contain mixed objects of different class and length
```{r}
list <- list(
  given_names = c("John", "Meriam", "Judith", "Cristine"),
  cities = c("New York", "Amsterdam", "Berlin"),
  lgl = as.logical(1:20),
  favorite_places = list("Empire State Building", "Tiffany", "Central Park", "National Museum of Natural History", "Metz", "Amsterdam Paradiso")
  
)
is.vector(list)
is.recursive(list)

```

## Loops (or also called 'iterations')
Apply a function to a vector

 - Because a dataframe is also a list you can use map on a dataframe
 - There are multiple map functions
 
 Basic 'map' template
 
 `map(YOUR_LIST, YOUR_FUNCTION)`
 
## Simple iteration examples 
```{r}
library(purrr)
library(repurrrsive)

sum(table$b)
# loop over the individual 'vectors' (columns) in a dataframe
map(table, class)
# loop over the individual 'vectors' (elements) in a list
map(list, length)
```

## Vectorised operations
R will recycle shorter objects over longer, and will tell you about it.
```{r, warning = TRUE}
x <- c(1:5)
y <- c(1:3)
z <- x - y 
z
```

# Back to our case data "pertussis"

## Exploratory graphs
```{r}
names(pertussis_data_tidy)
pertussis_data_tidy %>%
  ggplot(aes(x = year, y = annual_pertussis_cases)) +
  geom_point(aes(position = "jitter" ))
```

## Grouping, filtering and selecting data with `{dplyr}`
Let's calculate all cases for all countries, execept Bulgaria in the ninetees. Which countrie(s) has/have the highest number of cases? Which the lowest?
```{r}
pertussis_data_tidy$year[1] 

ninetees <- c("1991-01-01", "1992-01-01", "1993-01-01", "1994-01-01", "1995-01-01", "1996-01-01", "1997-01-01", "1998-01-01", "1999-01-01")

ninetees_summary <- pertussis_data_tidy %>%
  dplyr::select(country, year, annual_pertussis_cases) %>%
  mutate(year = as.character(year)) %>%
  dplyr::filter(year %in% ninetees) %>%
  dplyr::filter(country != "Bulgaria") %>%
  group_by(country) %>%
  summarise(all_cases_ninetees = sum(annual_pertussis_cases, na.rm = TRUE)) %>%
  arrange(desc(all_cases_ninetees))
```

## Nintees data
```{r}
ninetees_summary
tail(ninetees_summary)
```

## Exploratory Data Analysis
```{r, echo=FALSE}
# pertussis_data_tidy
pertussis_data_tidy %>%
  na.omit() %>%
  group_by(country) %>%
  ggplot(aes(x = year,
             y = annual_pertussis_cases)) +
  geom_line(aes(group = country)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme_bw()
```

## Transform y-scale to log10, `colour = country`
```{r, echo=FALSE}
pertussis_data_tidy %>%
  na.omit() %>%
  group_by(country) %>%
  ggplot(aes(x = year,
             y = log10(annual_pertussis_cases))) +
  geom_line(aes(group = country, colour = country))  +
  theme_bw() +
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
This is a messy graph indeed!!

## Adding a linear regression model for "Netherlands" 
To reduce the data in the graph we can look at individual countries.
```{r}
pertussis_data_tidy$country %>% as_factor() %>% levels()

## filter data for the Nethelands
nl_data <- pertussis_data_tidy %>%
  dplyr::filter(country == "Netherlands")
```

## Plot data only NL
```{r}
nl_data %>% 
  na.omit() %>% ## for sake of simplicity we remove the years that are missing
  ggplot(aes(x = year,
             y = annual_pertussis_cases)) + 
  geom_line() +
  geom_point(colour = "red") +
  ggtitle("The Netherlands; Whooping Cough cases")
```

## Plot linear model for NL
```{r}

nl_model <- lm(
  annual_pertussis_cases ~ year, 
  data = nl_data %>%
    na.omit()
)

## plot model for NL
plot_model <- nl_data %>%
  na.omit() %>%
  add_predictions(nl_model) %>%
  ggplot(aes(x = year, 
             y = pred)) +
  geom_line() +
  ggtitle("Linear trend") +
  geom_point(
    data = nl_data,
    aes(x = year, 
    y = annual_pertussis_cases),
    colour = "red") +
  geom_line(
    data = nl_data %>% na.omit, ## note the pipe to remove NA
    aes(x = year, 
        y = annual_pertussis_cases),
    colour = "red",
    linetype = "dashed"
)

```

## Show model plot
```{r}
plot_model
```

## Plot residuals
```{r}
plot_res <- nl_data %>%
  na.omit() %>%
  add_residuals(nl_model) %>% 
  ggplot(aes(year, resid)) + 
  geom_hline(yintercept = 0, colour = "white", size = 3) + 
  geom_line() + 
  ggtitle("Remaining pattern") 

```

## Plot residuals
```{r}
plot_res
```

## How 'good' is this model?
```{r}
## tidyverse innovation
nl_model %>% broom::tidy()
## classical R output
summary(nl_model)
```

## How can we apply this model to `every` country 
without doing the above `r pertussis_data_tidy$country %>% as_factor() %>% levels() %>% length()` times.

## Split the data by country and apply the model to each subset

 - In fact, each country is a dataframe
 - We can subset the original dataframe into seperate dataframes for each country 

## The classical way:
`split()` >> `apply()` >> `combine()`
```{r}
pertussis_data_tidy_split <- split(pertussis_data_tidy,
                                   pertussis_data_tidy$country)
head(pertussis_data_tidy_split[[1]], 2)
```
But there is a better way in the `{tidyverse}`
Here we combine `{tidyr}`, `{purrr}` and `{broom}`

## **Discuss with your neighbour**:

 - What could be the downside for the `split()` >> `apply()` >> `combine()` approach?

## Better way to keep track of your results and models

 1. Write functions that plot and model the data of each subset
 2. Group the data by the subsetting factor (in our case this could be `country` or even `year`
 3. Nest the grouped data into a dataframe (inside a list column)
 _Are you following?_
 4. Apply the functions to your list column (by iteration, using the `map()` functions)
 5. Store the results back into the dataframe as a new list column
 6. Explore the results inside one big dataframe, the way you are used to

## Creating a list column with each country as an element in the list 
We also remove the NAs for now.
```{r}
nested_pertussis <- pertussis_data_tidy %>%
  na.omit() %>%
  dplyr::select(country, year, annual_pertussis_cases) %>%
  group_by(country) %>%
    nest 

head(nested_pertussis, 2) ## you see the grouping by country
head(nested_pertussis$data[[1]], 2) ## you get the individual countries
```

## **Discuss with you neighbour**

 - How would you label the idividual elements of the list column (`$data`)?
 
## Linear model for each country
First we write a function that creates our linear model
```{r}
country_model_lm <- function(df){
  model <- lm(
    annual_pertussis_cases ~ year, 
    data = df)
  return(model)
}
```

# Iterate the model function over nested `$data` with `purrr::map()`
```{r}
models <- map(
  nested_pertussis$data, country_model_lm
  )
```

## Keep it togetheR
 - We have the models now
 - It is better to store them together with the data
 - Do that by using `dplyr::mutate()` 
 
## Create an additional list-column on the basis of an existing one
```{r}
nested_pertussis <- nested_pertussis %>%
  mutate(models_lm = map(data, country_model_lm))
```

## Adding residuals to all the models
```{r}
nested_pertussis <- nested_pertussis %>% 
  mutate(
    resids = map2(data, models_lm, add_residuals)
  )
```

## **Discuss with your neighbour**
 - What type of arguments does the function `map2()` take?
 - How would you find more information?

## Unnest the residuals
```{r}
resids <- nested_pertussis %>%
  unnest(resids)
```

## Plotting all residuals
```{r}
plot_all_resids <- resids %>% 
  ggplot(aes(year, resid)) +
    geom_line(aes(group = country, colour = country), alpha = 0.8) + 
    geom_smooth(se = FALSE) +
  theme_bw() +
  theme(legend.position="none") +
  facet_wrap( ~ country) +
 theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## What can we conclude from this graph?
```{r, fig.width=7, fig.height=5}
plot_all_resids
```

# Extracting information from a list column ***ADVANCED***

## Looking at quantative statistical measures for model quality
```{r}
r_squared <- nested_pertussis %>% 
  mutate(glance = map(models_lm, broom::glance)) %>% 
  unnest(glance, drop = TRUE) %>%
  select(country, r.squared, AIC, BIC) %>%
  arrange(r.squared)
```

## Plotting r.sqared values
```{r}
r_squared %>%
  ggplot(aes(x = reorder(as_factor(country), r.squared),
             y = r.squared)) +
  geom_point() +
  coord_flip() +
  ylab("R-squared") +
  xlab("Countries") + 
  geom_point(data = filter(r_squared, r.squared <= 0.25), colour = "red")
```

## Plotting pertussis cases for countries with low r.squared over time
 1. Filter countries with `r.squared` <= 0.25
 2. Put  countries in vector
 3. Plot data

## Step 1
```{r}
low_r_squared <- r_squared %>%
  dplyr::filter(r.squared <= 0.25) %>%
  dplyr::select(country) 
low_r_squared <- low_r_squared$country
```

## Step 2
```{r}
low_r_squared_nested <- nested_pertussis %>%
  dplyr::filter(country %in% low_r_squared) %>%
  select(country, data) %>%
  unnest()
```

## Step 3
```{r}
low_r_squared_nested %>%
  ggplot(aes(x =year, y = annual_pertussis_cases)) +
  geom_line(aes(group = country, color = country)) +
    theme_bw() +
  theme(legend.position="none") +
  facet_wrap( ~ country, scales = "free") +
 theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
What is happening with the pertussis vaccination grade over the past 8 years? 

## Storing ggplot2 objects in a list-column
 1. Create a function that makes the plot
 3. Test function on single dataframe
 2. Apply the function using `mutate()` and `map()` to all dataframes or models
 
```{r}
## dummy variable
df <- nested_pertussis$data[[1]]
plot_country <- function(df){
  
  df %>%
    ggplot(aes(x = year,
           y = annual_pertussis_cases)) +
    geom_line() +
    geom_smooth() +
    ylab("Annual cases")

}

## test function
plot_country(df = df)
```

## Apply plotting function to nested data
```{r}
nested_pertussis <- nested_pertussis %>%
  mutate(plots_cases_over_time = map(data, 
                                     plot_country))

```

## Add countries as names to nest-column
```{r}
names(nested_pertussis$plots_cases_over_time) <- 
  c(nested_pertussis$country)

nested_pertussis$plots_cases_over_time[1]
```

## Pull out "The Netherlands"
```{r}
nested_pertussis$plots_cases_over_time$Netherlands
```

## Plotting a panel of 4 graphs
```{r, echo=FALSE}
grid <- cowplot::plot_grid(
  nested_pertussis$plots_cases_over_time$Armenia, 
  nested_pertussis$plots_cases_over_time$Austria,
  nested_pertussis$plots_cases_over_time$Ukraine,
  nested_pertussis$plots_cases_over_time$Netherlands, 
  labels = c("Armenia", "Austria", "Ukraine", "Netherlands"),
  scale = 0.7, hjust = c(-2.5, -3, -2.7, -1.5))
grid
```

## Literature
For some background on the pattern we are seeing

 - https://www.scientificamerican.com/article/why-whooping-cough-vaccines-are-wearing-off/
 - http://outbreaknewstoday.com/pertussis-cases-up-significantly-in-the-eu-netherlands-and-uk-worst-hit-55315/


## Exploring many more models
Let's add a quadratic model in the mix. Assume we want to explore non-linear relationships in this dataset
```{r}
non_linear_model <- function(df, model_params){
  
  nl_model <- lm(
    annual_pertussis_cases ~ poly(year, 
                                  model_params),
                 data = df)
  
  return(nl_model)
}
```

## Creating a safe version of this function
```{r}
safe_non_linear <- purrr::safely(non_linear_model)
## apply test:
df = nested_pertussis$data[[1]]
test_non_linear <-  df %>% non_linear_model(df = .,
                                            model_params = 2)
```

## Test function on one country
```{r}
test_non_linear %>% broom::glance()
```

## Apply model to all countries 
We rerun the steps above to add this new model and new graphs to the nested dataframe

Add new model to the nested table
```{r}
nested_pertussis <- nested_pertussis %>%
  mutate(models_nl_2 = map(data, safe_non_linear, 
                          model_params = 2))
  
nested_pertussis$models_nl_2 <- transpose(nested_pertussis$models_nl_2)
```

## Set names to elements in the list-column
To be able to `pluck()` by name later
```{r}
names(nested_pertussis$models_nl_2$result) <- nested_pertussis$country
```

## Pluck results in new list-column
```{r}
nested_pertussis$models_nl_2$result

nested_pertussis <- nested_pertussis %>%
  mutate(statistics_nl = pluck(models_nl_2, "result"))
nested_pertussis$statistics_nl[[1]]

```

## Tidy the list-column with `{broom}`
```{r}
nested_pertussis <- nested_pertussis %>%
  mutate(parameters_nl = map(statistics_nl, glance))
```


## Looking at quantative statistical measures for model quality
```{r}
r_squared_nl <- nested_pertussis %>%
  select(country, parameters_nl) %>%
  unnest()
```

## Plotting r.sqared values
```{r, echo=FALSE}
r_squared %>%
  ggplot(aes(
    x = reorder(as_factor(country), r.squared),
    y = r.squared
  )) +
  geom_point() +
  ylab("Countries") +
  geom_point(data = filter(
    r_squared,
    r.squared <= 0.25
  ), colour = "red") +
  geom_point(
    data = r_squared_nl,
    aes(
      x = reorder(as_factor(country), r.squared),
      y = r.squared
    ), colour = "purple",
    shape = 21, size = 3
  ) +
  theme_bw() +
  xlab("Country") +
  ylab("R.squared") +
  coord_flip() +
  geom_point(data = r_squared_nl %>%
               dplyr::filter(country == "Ireland" |
                             country == "Belgium"), 
             colour = "blue", shape = 22, size = 5)



```

## Let's examine two models for two countries where the non-linear did and did not not improve the R.squared: Ireland (improved) and Belgium (not-improved)  
```{r}

x <- nested_pertussis %>%
  select(country,
         data, 
         models_lm,
         statistics_nl) %>%
  gather(models_lm:statistics_nl, key = "models", value = "model_params") %>%
 print()  
```

## Remove 'empty model'
```{r}
ind <- x$model_params == "NULL"
#ind <- x$data == "NULL"
x <- x[!ind, ]
```

## Add prediction-list column
```{r}
predictions <- x %>%
#  filter(country == "Czech Republic") %>%
  mutate(predictions = map2(data, model_params, add_predictions,
         var = "annual_pertussis_cases")) %>%
  filter(country == "Ireland" | 
         country == "Belgium") %>%
  select(country, data, predictions)

```

## Set names
```{r}

names(predictions$predictions) <- predictions$country
```

## Belgium
```{r}
nested_pertussis$plots_cases_over_time$Belgium +
  geom_line(data = predictions$predictions[[1]], colour = "red", size = 2) +
  
  geom_line(data = predictions$predictions[[3]], colour = "green", size = 2) 
```  

## Ireland
```{r}
nested_pertussis$plots_cases_over_time$Ireland+
  geom_line(data = predictions$predictions[[2]], colour = "red", size = 2) +
  
  geom_line(data = predictions$predictions[[4]], colour = "green", size = 2) 
```


## Disclaimer & Licence

_The work in this presentation has been adapted from many sources as mentioned in slide 'Credits'. For a full reference list see the 'REFERENCES.txt' document in the github repo 
https://www.github.com/usashogeschoolutrecht/surf_research_bootcamp_2018

The work presented here may be shared, remixed or adapted as long as the original references and the authors of this document are mentioned in the redistribution: CC BY-SA
