---
title: "Statistics; Checking Assumptions"
author: "Marc A.T. Teunis"
date: "January 26, 2018"
output: html_document
  
---
<style>

.footer {
    font-size: 16pt;
    color: Black;
    font-family: 'Helvetica'
    content: attr(label);
    position: absolute;
    margin-left: auto;
    margin-right: auto;
    left: 0;
    right: 0;
    text-align:center;
    bottom: 10px;
    line-height: 1.9;
    display: block;
}

</style>

<div class="footer" style="margin-top;font-size:120%;"> 
Intro to Statistics Using R - November 2017 </div>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5, fig.height = 3,
                      warning = FALSE,
                      error = FALSE,
                      message = FALSE)
```

```{r, root_1, include=FALSE}
## defines the root of the project for later use
require("rprojroot") || utils::install.packages("rprojroot")
library(rprojroot)
root <- find_root_file(criterion = is_rstudio_project)
```

## References and sources

This tutorial was inspired by the
R Code for Chapter 5 of:

Field, A. P., Miles, J. N. V., & Field, Z. C. (2012). 

`Discovering Statistics Using R: and Sex and Drugs and Rock 'N' Roll. `

London Sage
(c) 2011 Andy P. Field, Jeremy N. V. Miles & Zoe C. Field

## Assumptions

 - Each statistical test implies assumptions for the design, the structure, the distribution and the variance in the data
 - Selecting the wrong test for the wrong data will lead to over- or underestimation of statistical validity and probably to wrong conclusions  
 - Assumptions can be checked by creating graphs and perfoming tests, there can be confusion!
 - Before checking assumptions we do `Exploratory Data Analysis`
 
## The data

 - Data from the book "Discovering Statitics Using R" 
 - Case: Dataset supplied by Sacha Kofman (Enzazaden NL)

## Install and Load Packages
```{r, packages}
# install.packages("car")
# install.packages("ggplot2")
# install.packages("pastecs")
# install.packages("psych")
library(car)
library(ggplot2)
library(pastecs)
library(psych)
# library(Rcmdr)
library(tidyverse)
```

## Read in the download data:
```{r}
dlf <- read_delim(file = file.path(root, 
                                   "data", 
                                   "DownloadFestival.dat"), 
                                   delim =  "\t", na = c("", " "))
dlf %>% head(3)
```

## Checking distribution and detecting outliers
```{r}
sum(is.na(dlf))
x <- summary(dlf)
min_maxs <- x[c(1, 6), c(3:5)] %>% unlist() %>% print()
```

## Detecting an outlier with a histogram
```{r}
hist.outlier <- ggplot(dlf, aes(day1)) + 
  geom_histogram(aes(y=..density..), 
                 colour="black", 
                 fill="white") + 
  labs(x="Hygiene score on day 1", y = "Density") +
  theme(legend.position = "none")
hist.outlier
```

## Transform data to a different format (convert to `long/stacked`)
```{r}
dlf_long <- dlf %>% 
  tidyr::gather(day1:day3, key = "days", value = "hygiene_score")
```

## Boxplots with outlier
```{r, echo=FALSE}
hist.boxplot <- dlf_long %>%
  ggplot(aes(x = days, y = hygiene_score)) + 
  geom_boxplot(aes(group = days)) +
  geom_point(data = dplyr::filter(dlf_long, hygiene_score > 19), 
             colour = "darkred", size = 2.5) +
  labs(x="Hygiene score on day 1", y = "Hygiene Score") +
  theme(legend.position = "none") + 
  facet_wrap(~ gender)
hist.boxplot
```

## Remove outlier
```{r}
dlf <- dlf %>%
  dplyr::filter(!day1 > 19)
```

## Boxplots without outlier
```{r}
hist.boxplot <- dlf %>%
  tidyr::gather(day1:day3, key = "days", value = "hygiene_score") %>%
  ggplot(aes(x = days, y = hygiene_score)) + 
  geom_boxplot(aes(group = days)) + 
  labs(x="Hygiene score on day 1", y = "Hygiene Score") +
  theme(legend.position = "none") + 
  facet_wrap(~ gender)
hist.boxplot
```

# Histograms for hygiene scores on day 1, day 2 and day 3.

## Histogram for day 1:
```{r}
hist.day1 <- ggplot(dlf, aes(day1)) + 
  geom_histogram(aes(y=..density..), 
                 colour="black", 
                 fill="white") + 
  labs(x="Hygiene score on day 1", y = "Density") +
  theme(legend.position = "none")
  
hist.day1
```

## Histogram for day 2
```{r}
hist(dlf$day2)

hist.day2 <- ggplot(dlf, aes(day2)) + 
  geom_histogram(aes(y=..density..), 
                 colour="black", 
                 fill="white") + 
  labs(x="Hygiene score on day 2", y = "Density") +
  theme(legend.position = "none")
hist.day2
```

## Histogram for day 3
```{r}
hist.day3 <- ggplot(dlf, aes(day3)) + 
  theme(legend.position = "none") + 
  geom_histogram(aes(y=..density..), 
                 colour="black", fill="white") + 
  labs(x="Hygiene score on day 3", y = "Density")
hist.day3
```

```{r, include=FALSE}
## store to disk
imageDirectory <- file.path(root, "images")
p1 <- hist.day1 + 
  stat_function(fun = dnorm, 
                args = list(mean = mean(dlf$day1, na.rm = TRUE), 
                sd = sd(dlf$day1, na.rm = TRUE)), 
                colour = "black", size = 1)

ggsave(file = paste(imageDirectory,"05 DLF Day 1 Hist.png",sep="/"))

p2 <- hist.day2 + 
  stat_function(fun = dnorm, 
                args = list(mean = mean(dlf$day2, 
                na.rm = TRUE), 
                sd = sd(dlf$day2, na.rm = TRUE)), 
                colour = "black", size = 1)

ggsave(file = paste(imageDirectory,"05 DLF Day 2 Hist.png",sep="/"))

p3 <- hist.day3 + 
  stat_function(fun = dnorm, 
                args = list(mean = mean(dlf$day3, 
                na.rm = TRUE), 
                sd = sd(dlf$day3, na.rm = TRUE)), 
                colour = "black", size = 1) +
  

ggsave(file = paste(imageDirectory,"05 DLF Day 3 Hist.png",sep="/"))
```

## Panel of histograms + distribution density
```{r, fig.width=6, fig.height=4}
cowplot::plot_grid(p1, p2, p3, nrow = 1)
```

# Check for normality by graphs; qq-plot

## Q-Q plot for day 1
```{r}
qqplot.day1 <- qplot(sample = dlf$day1, geom = "qq")
qqplot.day1
```

## Or antoher way with a home brewed function in ggplot2
```{r}
source(file = file.path(root, "code", "ggqq.R"))
gg_qq_1 <- gg_qq(dlf$day1)
gg_qq_1
```

## Q-Q plot for day 2:
```{r}
qqplot.day2 <- qplot(sample = dlf$day2, geom = "qq")
qqplot.day2
```

## Or via the other alternative way
```{r}
gg_qq_2 <- gg_qq(dlf$day2)
gg_qq_2
```

## Q-Q plot of the hygiene scores on day 3
```{r}
qqplot.day3 <- qplot(sample = dlf$day3, geom = "qq")
qqplot.day3
```

## Or
```{r}
gg_qq(dlf$day3)
```

## Skewness and kurtosis
Skewness and kurtosis are parameters that display the deviation from normality looking at the shape of the distribution polynom. A distribution with an absolute `skew.2SE` > 1 is significantly skewed and not normal. A distribution with an absolute `kurt.2SE` > 1 has significant kurtosis and is not normally distributed

`kurt.2SE` and `skew.2SE` are calculated from

 - $kurt.2SE = kurt / 2*(standard.error)$ 
 - $skew.2SE = skew / 2*(standard.error)$
 
## `Shapiro-Wilk test` 
To test for normality we can use the `Shapiro-Wilk test`. This test checks whether the deviation from normality is significant (H0) or not (H1), 

 - p-value < 0.05 means that the distribution is significantly different from a normal distribution: assumption "the distribution is not normal"
 - p-value > 0.05 means that the distribution is not significantly different from normal: assumption "the distribution is normal"
 
## Quantifying normality with numbers
```{r}
library(psych)		
describe(dlf$day1)
```

## Two alternative ways to describe multiple variables.
```{r}
describe(cbind(dlf$day1, dlf$day2, dlf$day3)) %>% knitr::kable()
```

## Second way
```{r}
knitr::kable(describe(dlf[,c("day1", "day2", "day3")]))
```


## `stat.desc()` to get descriptive statistics for a dataframe or a variable 
```{r}
library(pastecs)
stat.desc(dlf$day1, basic = FALSE, norm = TRUE)
stat.desc(cbind(dlf$day1, dlf$day2, dlf$day3), basic = FALSE, norm = TRUE)
round(stat.desc(dlf[, c("day1", "day2", "day3")], basic = FALSE, norm = TRUE), digits = 3)
```

# R exam case

The R exam data file `./data/RExam.dat` contains data on the performance of students on an R exam (data from Field et al., 2012).
The variables:

This data consists of 5 variables: 

 - `exam` first-year exams scores as a percentage (numeric / continous)
 - `computer` measure of computer literacy as a percentage (numeric / continuous)
 - `lectures` percentage of lectures in the R class attended, as a percentage (numeric / continuous)
 - `numerical` arbitrary score from 1-15 on the ability to cope with numbers ("statistical literacy") (integer / ordinal)
 - `uni` the university where the student taking the R class attends (factor)

## Read in R exam data.
```{r}
rexam <- read_delim(file = file.path(root, "data", "RExam.dat"), delim = "\t")
rexam %>% head(3)
```

## Set the variable `uni` to be a factor
```{r}
library(forcats)
rexam <- rexam %>%
  mutate(uni = as.factor(uni)) %>%
  mutate(uni = forcats::fct_recode(uni, duncetown_university = "0", sussex_university = "1")) 
rexam %>% head(3)
```

## Long format
```{r}
rexam_long <- rexam %>%
  gather(exam:numeracy, key = "measure", value = "value")
rexam_long
```

## Normality test and skewness
```{r}
round(stat.desc(rexam[, c("exam", "computer", "lectures", "numeracy")], 
                basic = FALSE, norm = TRUE), 
                digits = 3)
```

# Histograms of all variables

## Histogram for all data together
```{r}
all_data_hist <- rexam_long %>%
  ggplot(aes(value)) + 
  geom_histogram(aes(y=..density..), 
                colour="black", fill="white")
all_data_hist
```

## `exam` histogram
```{r}
hexam <- ggplot(rexam, aes(exam)) + 
  geom_histogram(aes(y=..density..), 
                colour="black", fill="white") + 
  labs(x = "First Year Exam Score", 
                y = "Density") + 
  stat_function(fun = dnorm, 
                args = list(mean = mean(rexam$exam, na.rm = TRUE), 
                sd = sd(rexam$exam, na.rm = TRUE)), 
                colour = "red", size = 1) + 
  theme(legend.position = "none") 
hexam
```

## `computer` histogram
```{r, eval=FALSE}
hcomputer <- ggplot(rexam, aes(computer))  + 
  geom_histogram(aes(y=..density..), 
                 colour="black", 
                 fill="white") + 
  labs(x = "Computer Literacy", 
                 y = "Density") + 
  stat_function(fun = dnorm, 
                 args = list(mean = mean(rexam$computer, na.rm = TRUE), 
                 sd = sd(rexam$computer, na.rm = TRUE)), 
                 colour = "red", size = 1) + 
  theme(legend.position = "none")
hcomputer
```

## `lectures`
```{r}
hlectures <- ggplot(rexam, aes(lectures)) + 
  geom_histogram(aes(y=..density..), 
                 colour="black", fill="white") + 
  labs(x = "Percentage of Lectures Attended", y = "Density") + 
  stat_function(fun = dnorm, 
                 args = list(mean = mean(rexam$lectures, na.rm = TRUE), 
                 sd = sd(rexam$lectures, na.rm = TRUE)), 
                 colour = "red", size = 1) + 
  theme(legend.position = "none")
hlectures
```

## `numeracy`
```{r}
hnumeracy <- ggplot(rexam, aes(numeracy)) + 
  geom_histogram(aes(y=..density..), 
                 colour="black", fill="white") + 
  labs(x = "Numeracy", 
                 y = "Density") + 
  stat_function(fun = dnorm, 
                 args = list(mean = mean(rexam$numeracy, na.rm = TRUE), 
                 sd = sd(rexam$numeracy, na.rm = TRUE)), 
                 colour = "red", size = 1) + 
  theme(legend.position = "none")
hnumeracy
```

# Using `{purrr}` to get statistics for many variables
```{r}
library(broom)
library(purrr)
```

## `nest()` to get a list of dataframes
```{r}
names(rexam)
rexam_by_var <- rexam %>%
  as_tibble() %>%
  gather(exam:numeracy, key = "measure", value = "value") %>%
  group_by(uni, measure) %>%
  nest()
rexam_by_var
```

## Adding a column with descriptive statistics with `mutate()`
```{r}  
rexam_descr <- rexam_by_var %>%
  mutate(descriptives = map(data, stat.desc, basic = TRUE, norm = TRUE))
rexam_descr
```

## Shapiro: 

## Getting the Shapiro-Wilk results: a function
```{r}
get_shapiro_wilk <- function(df){
  p_value <- df[ , 1][20]
  return(p_value)
}
```

## Apply function to nested table and unnest
```{r}
rexam_descr <- rexam_descr %>%
  mutate(shap_wilk = map(descriptives, get_shapiro_wilk))
descriptives <- unnest(rexam_descr, shap_wilk)
descriptives[ , c("uni", "measure", "shap_wilk")]
```

## Plot results Shapiro-Wilk test
```{r}
descriptives[ , c("uni", "measure", "shap_wilk")] %>%
  ggplot(aes(x = measure, y = shap_wilk)) +
  geom_point(aes(colour = uni), size = 2) +
  geom_hline(yintercept= 0.05, linetype="dotted", size = 1.5)

```

## Plotting a histrogram for each `measure` and each university, using `facets`
```{r}
hist_measure <- rexam_long %>%
    ggplot(aes(x = value)) +
  geom_histogram(aes(y = ..density..), 
                 fill = "white", 
                 colour = "black", 
                 binwidth = 1) + 
  labs(x = "Arbitrary Units", y = "Density") +
  facet_grid(uni ~ measure) 
#hist_measure
```

## Plot `measure`
```{r}
hist_measure
```

## Plotting a histogram per `computer` using `dplyr::filter()`; `computer`
```{r}
 hist_computer <- rexam_long %>%
  dplyr::filter(measure == "computer") %>%
    ggplot(aes(x = value)) +
  geom_histogram(aes(y = ..density..), 
                 fill = "white", 
                 colour = "black", 
                 binwidth = 1) + 
  labs(x = "Arbitrary Units", y = "Density") + 
  ggtitle("Computer literacy") +
  geom_line(aes(y = ..density..), colour=26, stat = 'density', size = 2, alpha = .6) +
  facet_wrap(~ uni)
```

## Plot `computer`
```{r}
hist_computer
```

## Plotting a histogram per group using `dplyr::filter()`; `exam`
```{r}
plot_exam <- rexam_long %>%
  dplyr::filter(measure == "exam") %>%
    ggplot(aes(x = value)) +
  geom_histogram(aes(y = ..density..), 
                 fill = "white", 
                 colour = "black", 
                 binwidth = 1) + 
  labs(x = "Arbitrary Units", y = "Density") + 
  ggtitle("First year exam") +
  geom_line(aes(y = ..density..), colour=25, stat = 'density', size = 2, alpha = .6) +
  facet_wrap(~ uni)
```

## Plot `exam`
```{r}
plot_exam
```

## Plotting a histogram per group using `dplyr::filter()`; `numeracy`
```{r}
plot_numeracy <- rexam_long %>%
  dplyr::filter(measure == "numeracy") %>%
    ggplot(aes(x = value)) +
  geom_histogram(aes(y = ..density..), 
                 fill = "white", 
                 colour = "black", 
                 binwidth = 1) + 
  labs(x = "Arbitrary Units", y = "Density") + 
  ggtitle("Numerical literacy") +
  geom_line(aes(y = ..density..), colour=26, stat = 'density', size = 2, alpha = .6) +
  facet_wrap(~ uni)
```

## Plot `numeracy`
```{r}
plot_numeracy
```

## qqplots for the two variables
```{r}
qplot(sample = rexam$exam, stat="qq")
qplot(sample = rexam$numeracy, stat="qq")
```

## Testing for equal variance (sample estimates)

$s_2 = {\frac{1}{{N - 1}}\sum\limits_{i = 1}^N {\left( {x_i - \bar x} \right)^2 } }$

$sd = {\sqrt\frac{1}{{N - 1}}\sum\limits_{i = 1}^N {\left( {x_i - \bar x} \right)^2 } }$

## Levene's Test
```{r, eval=FALSE}

leveneTest(rexam$exam, rexam$uni)
leveneTest(rexam$exam, rexam$uni, center = mean)
leveneTest(rexam$numeracy, rexam$uni)
```

# Fixing data errors and data transformations

## Transforming data

 - To remove skewness or kurtosis
 - Apply the __*same*__ transformation to __*all variables*__
 - After transformation and analysis, sometimes you need to inverse-transform
 - It can be a time comsuming process: 'trial-and-error'

--> Exercise 


