---
title: "Exercise - Testing Assumptions"
author: "Marc A.T. Teunis"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      error = FALSE,
                      results = 'hide',
                      fig.show = 'hide')
```

```{r, root_1, include=FALSE}
## defines the root of the project for later use
require("rprojroot") || utils::install.packages("rprojroot")
library(rprojroot)
root <- find_root_file(criterion = is_rstudio_project)
```

```{r, packages}
library(tidyverse)

```

__Write an Rmd file, containing code chunks (if applicable) and answer all the questions below. Store the file in a folder called: "./answers_exercises" in your course project folder. Give this file the same name as the title of the current Rmd file__

## Data Transformations

1A) Load the `DownloadFestival(No Outlier).dat` data into R from the folder "./data"

```{r}
dlf <- read_delim(file = file.path(root, 
                  "data",                                                                   "DownloadFestival(No Outlier).dat"),
                  delim = "\t", 
                  na = c("", " "))
dlf %>% head(3)
```

 - Create a 'long' format of the data, gather the Day1:Day3 variables to achieve this, call your new variable for the days `day` and the values you call `hygiene_score`. Use `tidyr::gather` to get there.
 - Transform the `score` variable by adding four new variables: `score_log10` (log10 transform), `score_log2` (log2 transform), `score_power2` (take each score to the power 2), `score_div_mean` (divide each hygiene score through the mean for the all hygiene scores). 
 - In this last transformation, how would you deal with missing values?

```{r}
# long format
dlf_long <- dlf %>% 
  tidyr::gather(day1:day3, key = "days", value = "hygiene_score")

# transform

mean_hygiene_score <- mean(dlf_long$hygiene_score, na.rm = TRUE)

dlf_long <- dlf_long %>%
  dplyr::mutate(score_log10 = log10(hygiene_score),
                score_log2 = log2(hygiene_score),
                score_power2 = hygiene_score^2,
                score_div_mean = hygiene_score / mean_hygiene_score)


```




1B) Plot histograms for `score`, `score_log10`, `score_log2`, `score_div_mean`. Facet the histograms for each `day`. Use `ggplot2` to achieve you goal.

```{r}
# untransformed scores
names(dlf_long)
hist_scores <- ggplot(dlf_long, aes(hygiene_score)) + 
  geom_histogram(aes(y=..density..), 
                 colour="black", 
                 fill="white") + 
  labs(x="Hygiene score on day 1", y = "Density") +
  theme(legend.position = "none") +
  facet_grid(gender ~ days)
hist_scores

# log 10 transformed scores
hist_scores_log10 <- ggplot(dlf_long, aes(score_log10)) + 
  geom_histogram(aes(y=..density..), 
                 colour="black", 
                 fill="white") + 
  labs(x="Hygiene score on day 1", y = "Density") +
  theme(legend.position = "none") +
  facet_grid(gender ~ days)
hist_scores_log10

# log2 transfromed
hist_scores_log2 <- ggplot(dlf_long, aes(score_log2)) + 
  geom_histogram(aes(y=..density..), 
                 colour="black", 
                 fill="white") + 
  labs(x="Hygiene score on day 1", y = "Density") +
  theme(legend.position = "none") +
  facet_grid(gender ~ days)
hist_scores_log2

hist_scores_power2 <- ggplot(dlf_long, aes(score_power2)) + 
  geom_histogram(aes(y=..density..), 
                 colour="black", 
                 fill="white") + 
  labs(x="Hygiene score on day 1", y = "Density") +
  theme(legend.position = "none") +
  facet_grid(gender ~ days)
hist_scores_power2

ggsave(filename = file.path(root, "images", "some_silly_histogram.svg"))

ggsave(filename = file.path(root, "images", "some_silly_histogram.png"))


```

1C) Which transformation worked best? Do you know of a way to fomally test this?
If so, apply your test to the data

```{r, eval=FALSE}
## Check with Shapiro Wilk: see presentations for Code
library(pastecs)
names(dlf_long)

shapiro.test(dlf$day1) # 
shapiro.test(dlf$day2) 
shapiro.test(dlf$day3)



hist(dlf$day1)


dlf_nested <- dlf_long %>%
  group_by(days, gender) %>%
  nest()

item_1_nested <- dlf_nested$data[[1]]

stat.desc(item_1_nested, norm = TRUE)

dlf_nested$data[[1]]

dlf_descr <- dlf_nested %>%
  mutate(descriptives = map(data, stat.desc, basic = TRUE, norm = TRUE))


dlf_descr$descriptives[[1]][20, c("hygiene_score", "score_log10")]

get_shapiro_wilk <- function(df){
  p_value <- df[20 , c("hygiene_score", 
                       "score_log10",
                       "score_log2",
                       "score_power2",
                       "score_div_mean")]
  return(p_value)
}

## apply get shapiro to nested table
dlf_descr <- dlf_descr %>%
  mutate(shap_wilk = map(descriptives, get_shapiro_wilk))

descriptives <- unnest(dlf_descr, shap_wilk)
descriptives[ , c("days", 
                  "gender", 
                  "hygiene_score", 
                  "score_log10",
                  "score_log2",
                  "score_power2",
                  "score_div_mean")]


class(dlf_nested$data)

```

## Example data: Enzazaden, 

Use the dataset `Wateruptake.xlsx` in folder "./data"

2A) Load the data into R using the `{readxl}` package
```{r}
library(readxl)
library(tidyverse)
enza <- readxl::read_xlsx(path = file.path(root, "data", "Wateruptake.xlsx"))

new_names <- enza[1,] %>% as.character
new_names[1] <- "date_time"

names(enza) <- new_names
enza <- enza[-1,]

names(enza) <- str_replace_all(names(enza), 
                               pattern = " ", 
                               replacement = "_") %>%
tolower()


names(enza)
enza_long <- enza %>%
  gather(tomaat_1:prei_3, key = "vegetable", value = "wateruptake")

enza_long$wateruptake <- as.numeric(enza_long$wateruptake)
enza_long$vegetable <- as.factor(enza_long$vegetable)


enza_long

```

2B) Inspect the data, do you need to clean the data?
```{r}
## many NAs
sum(is.na(enza_long))

## check class of vars
lapply(enza_long[, c(1:ncol(enza_long))], class)
```

2C) Create a tidy dataframe:

 - lower_case variable names
 - long format of the data, if not already
 - convert to factors if necessary

```
# see above
```

2D) Check normality with Shapiro and Perform a Levene Test for equal variance
```{r}

## add vegetable type
enza_long <- enza_long %>%
  separate(vegetable, 
  into = c("vegetable_type", "id"), sep = "_", remove = FALSE)

## normality plots

source(file = file.path(root, "code", "ggqq.R"))

gg_qq_1 <- gg_qq(enza_long$wateruptake)
gg_qq_1 <- gg_qq(log10(enza_long$wateruptake)) 
  
# histograms
hist_enza_all <- ggplot(enza_long, aes(wateruptake)) + 
  geom_histogram(aes(y=..density..), 
                 colour="black", 
                 fill="white") + 
  labs(x="wateruptake", y = "Density") +
  theme(legend.position = "none") 
hist_enza_all

hist_enza <- ggplot(enza_long, aes(wateruptake)) + 
  geom_histogram(aes(y=..density..), 
                 colour="black", 
                 fill="white") + 
  labs(x="wateruptake", y = "Density") +
  theme(legend.position = "none") +
  facet_wrap(~ vegetable_type)
hist_enza

hist_enza_log10 <- ggplot(enza_long, aes(log10(wateruptake))) + 
  geom_histogram(aes(y=..density..), 
                 colour="black", 
                 fill="white") + 
  labs(x="wateruptake", y = "Density") +
  theme(legend.position = "none") +
  facet_wrap(~ vegetable_type)
hist_enza_log10


```


```{r}
## Shap-Wilk / Levene


enza_nested <- enza_long %>%
  group_by(vegetable_type) %>%
  nest()

library(pastecs)
library(car)
names(enza_long)

enza_nested$data[[1]]

enza_descr <- enza_nested %>%
  mutate(descriptives = map(data, stat.desc, basic = TRUE, norm = TRUE))


enza_descr$descriptives[[1]][20, ]

get_shapiro_wilk <- function(df){
  p_value <- df[20 , c("wateruptake")]
  return(p_value)
}

## apply get shapiro to nested table
enza_descr <- enza_descr %>%
  mutate(shap_wilk = map(descriptives, get_shapiro_wilk))

enza_descr$shap_wilk

descriptives <- unnest(enza_descr, shap_wilk)
descriptives[, "shap_wilk"]




## isolate only complete cases
ind <- complete.cases(enza)

enza_complete_long <- enza[ind, ] %>% 
  gather(tomaat_1:prei_3, 
         key = "vegetable", 
         value = "wateruptake") %>%                
  separate(vegetable, 
  into = c("vegetable_type", "id"), 
  sep = "_", remove = FALSE) %>%
  mutate(wateruptake = as.numeric(wateruptake))


hist_enza_all_complete <- ggplot(enza_complete_long, aes(wateruptake)) + 
  geom_histogram(aes(y=..density..), 
                 colour="black", 
                 fill="white") + 
  labs(x="wateruptake", y = "Density") +
  theme(legend.position = "none") 
hist_enza_all_complete


enza_nested_complete <- enza_complete_long %>%
    separate(vegetable, 
  into = c("vegetable_type", "id"), sep = "_", remove = FALSE) %>%
  mutate(wateruptake = as.numeric(wateruptake)) %>%
  group_by(vegetable_type) %>%
  nest()

tomato_complete <- enza_nested_complete$data[[1]][, "wateruptake"] %>% na.omit
onion_complete <- enza_nested_complete$data[[2]][, "wateruptake"] %>% na.omit
leek_complete <- enza_nested_complete$data[[3]][, "wateruptake"] %>% na.omit

leveneTest(wateruptake ~ vegetable_type, data = enza_long)
leveneTest(wateruptake ~ vegetable_type, data = enza_complete_long)


```

3) How could you fix the normality problem?
