---
title: "Statistics; Repeated measures ANOVA (GLM4)"
author: "Marc A.T. Teunis"
date: "`r Sys.Date()`"
output:
  html_document:
  ioslides_presentation: 
    highlight: pygments
    widescreen: yes
---
<style>

.footer {
    font-size: 16pt;
    color: Black;
    font-family: 'Helvetica'
    content: attr(label);
    position: absolute;
    margin-left: auto;
    margin-right: auto;
    left: 0;
    right: 0;
    text-align:center;
    bottom: 10px;
    line-height: 1.9;
    display: block;
}

</style>

<div class="footer" style="margin-top;font-size:120%;"> 
Data Analysis Using R - November 2017 </div>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 10, fig.height = 8,
                      warning = FALSE,
                      error = FALSE,
                      message = FALSE)
```

```{r, root_1, include=FALSE}
## defines the root of the project for later use
require("rprojroot") || utils::install.packages("rprojroot")
library(rprojroot)
root <- find_root_file(criterion = is_rstudio_project)
```

## References and sources

R Code for Chapter 13 of:
Field, A. P., Miles, J. N. V., & Field, Z. C. (2012). Discovering Statistics Using R: and Sex and Drugs and Rock 'N' Roll. 
London Sage
(c) 2011 Andy P. Field, Jeremy N. V. Miles & Zoe C. Field

## Install packages
```{r}
# install.packages("compute.es")
# install.packages("ez")
# install.packages("ggplot2")
# install.packages("multcomp")
# install.packages("nlme")
# install.packages("pastecs")
# install.packages("reshape")
# install.packages("WRS", repos="http://R-Forge.R-project.org")
# install.packages("akima")

#Initiate packages
library(compute.es)
library(ez)
library(ggplot2)
library(multcomp)
library(nlme)
library(pastecs)
library(reshape)
library(WRS)
## source("http://www-rcf.usc.edu/~rwilcox/Rallfun-v14")
# install.packages("nlme")
```

## Assumptions
From the testing assumptions we know that there is a normality problem and that there are many missing values. For the analysis below we disregard the normality issue for now. We assume that the data is normally distributed

## Vegetables example
```{r}
library(readxl)
library(tidyverse)
enza <- readxl::read_xlsx(path = file.path(root, "data", "Wateruptake.xlsx"))

new_names <- enza[1,] %>% as.character
new_names[1] <- "date_time"

names(enza) <- new_names
enza <- enza[-1,]

names(enza) <- str_replace_all(names(enza), 
                               pattern = " ", 
                               replacement = "_") %>%
tolower()


names(enza)
enza_long <- enza %>%
  gather(tomaat_1:prei_3, key = "vegetable", value = "wateruptake")

enza_long$wateruptake <- as.numeric(enza_long$wateruptake)
enza_long$vegetable <- as.factor(enza_long$vegetable)


enza_long

```

How man(y rows does the dataframe have?
`r nrow(enzo_long)`

## Scatterplot
```{r}
enza_long %>% ggplot(aes(x = date_time, y = wateruptake)) +
  geom_point(aes(colour = vegetable))


```

## Trimming could solve normality
```{r}
hist_enza_all <- ggplot(enza_long, aes(wateruptake)) + 
  geom_histogram(aes(y=..density..), 
                 colour="black", 
                 fill="white") + 
  labs(x="wateruptake", y = "Density") +
  theme(legend.position = "none") 
hist_enza_all

## keep all data with wateruptake < 12.5

enza_trimmed <- enza_long %>%
  filter(wateruptake < 10)

hist_enza_all_trimmed <- ggplot(enza_trimmed, aes(wateruptake^2)) + 
  geom_histogram(aes(y=..density..), 
                 colour="black", 
                 fill="white") + 
  labs(x="wateruptake", y = "Density") +
  theme(legend.position = "none") 
hist_enza_all_trimmed

shapiro.test(enza_trimmed$wateruptake)

```

Did we solve the normality issue?


## EDA
```{r}
enza_long %>% ggplot(aes(x = date_time, y = wateruptake)) +
  geom_point(aes(colour = vegetable))
```

## Anova with `aov`
```{r}
enza_aov <- aov(wateruptake ~ date_time, data = enza_long)
summary(enza_aov)
```

## Adding an extra variable `vegatable_type`
```{r}
names(enza_long)

# enza_long$vegetable <- as.character(enza_long$vegetable)

enza_long <- enza_long %>%
  separate(vegetable, 
  into = c("vegetable_type", "id"), sep = "_", remove = FALSE)

names(enza_long)
enza_long
```

## Two way ANOVA with `time` and `vegetable_type` and their interaction `time:vegetable_type` as independent factors (non-repeated measures)
```{r}
enza_long$date_time <- as.factor(enza_long$date_time)
aov_two_way_enza <- aov(wateruptake ~ date_time + vegetable_type + 
                        date_time:vegetable_type, 
                        data = enza_long)
summary(aov_two_way_enza)
```


## Setting and using contrasts
```{r}
enza_long$vegetable_type <- as.factor(enza_long$vegetable_type)

levels(enza_long$vegetable_type %>% na.omit)

contrasts <- cbind(c(0, 1, -1),
                  c(1, -1, 0))


aov_two_way_enza <- aov(wateruptake ~ vegetable_type,
                        data = enza_long)
                        

pairwise.t.test(enza_long$wateruptake, enza_long$vegetable_type,
                p.adjust.method = "BH")

```

## Interpretation of ANOVA results
This independent ANOVA with `aov()` reveals a significant effect of `vegetable_type` on `wateruptake`, meaning that the wateruptake between vegetable types in the dataset differs significantly. The effect of `date_time` on `wateruptake` is also significant, meaning that wateruptake increases (see the graph) significantly over time. There is no significant interaction, meaning that the effect that time has on wateruptake is the same for every vegetable type (the lines run parallel or do not cross). 

Let's look a summary plot for the lines to see if they indeed run parallel:
```{r}
library(lubridate)
enza_long$date_time <- as_datetime(enza_long$date_time)
enza_long %>%
  group_by(vegetable_type, date_time) %>%
  summarise(mean_wateruptake = mean(wateruptake)) %>%
  ggplot(aes(x = lubridate::as_datetime(date_time),
             y = mean_wateruptake)) +
  geom_point(aes(colour = vegetable_type), size = 4) +
  geom_point(data = enza_long, aes(x = lubridate::as_datetime(date_time), y = wateruptake, colour = vegetable_type), size = 0.2, alpha = 4/5)

```

What do you think?

 - What - Is the mean a good model for the data?
 - How could you deal with the missing values?
 other model would suit this data?
 
## Let's transform the `wateruptake`
```{r}
library(lubridate)
enza_long$date_time <- as_datetime(enza_long$date_time)
enza_long %>%
  group_by(vegetable_type, date_time) %>%
  summarise(mean_wateruptake = mean(wateruptake)) %>%
  ggplot(aes(x = lubridate::as_datetime(date_time),
             y = mean_wateruptake^8)) +
  geom_point(aes(colour = vegetable_type), size = 4) +
  geom_point(data = enza_long, aes(x = lubridate::as_datetime(date_time), y = wateruptake^8, colour = vegetable_type), size = 0.2, alpha = 4/5) 
```
 
## Transformations
So we could conclude that taken the i-th power of each value for wateruptake yield an approximate straight line. Is this a good idea? 

## Checking normality
Do you know how to check normality of this data?

## Repeated measures ANOVA using `lme`
First we remove the onions because the data for this vegetable is very incomplete and possible disrupts the model too much. 
```{r}

sum(is.na(enza_long))
enza_long_no_na <- na.omit(enza_long)


removed_onion <- enza_long %>%
  filter(vegetable_type != "ui") %>%
  droplevels() %>%
  mutate(date_time = as.factor(date_time),
         vegetable_type = as_factor(vegetable_type),
         wateruptake = scale(.$wateruptake, scale = FALSE)) %>%
  na.omit()


```


```{r, eval = FALSE}
library(tidyverse)

## baseline model
veggies_baseline <-lme(wateruptake ~ 1, 
                         random = ~1|vegetable/date_time, 
                         data = removed_onion, method = "ML")


anova(veggies_baseline) %>% stargazer::stargazer()


## only fixed effects
veggies_fixed <- lm(wateruptake ~ vegetable_type, data = removed_onion)
AIC(veggies_fixed)
AIC(veggies_baseline) # select the model with lowest AIC (Akaike Information Criteria)


## add vegetable_type as predictor
veggies_veggie_type <-update(veggies_baseline, .~. + vegetable_type)
AIC(veggies_veggie_type)
anova(veggies_veggie_type)

## add date_time as predictor
veggies_date_time <-update(veggies_veggie_type, .~. + date_time)
AIC(veggies_date_time)
anova(veggies_date_time)

## add interaction
# veggies_full <-update(veggies_date_time, .~. + vegetable_type:date_time)
# does not work due to numerical instability of the data:
# Error in MEEM(object, conLin, control$niterEM) : 
#  Singularity in backsolve at level 0, block 1
# try transformations!!

write_rds(veggies_date_time, path = file.path(root, "data", "mlm_complete_model.rds"))

results_multilevel_model <-anova(veggies_baseline, veggies_veggie_type, 
                                 veggies_date_time)
results_multilevel_model
summary(veggies_date_time)

library(readr)

write_rds(results_multilevel_model, path = file.path(root, "data", "wateruptake_model.rds"))

```

# Explore data - some more!

## Impute or extrapolate missing values
```{r}

anova_models <- read_rds(path = file.path(root, "data", "wateruptake_model.rds"))
complete_mlm <- read_rds(path = file.path(root, "data", "mlm_complete_model.rds"))

## missingness
pMiss <- function(x){sum(is.na(x))/length(x)*100}
apply(enza,2,pMiss)
apply(enza,1,pMiss)

## amelia to plot missing values
# install.packages("Amelia")
library(Amelia)
missmap(enza, main = "Missing values vs observed")


## There is a lot of missingness in the vegetables
## We can try to extrapolate the data from the point where it is missing on the basis of a logistic regression model
enza_long$date_time <- as_datetime(enza_long$date_time)
enza_long %>%
  group_by(vegetable_type, date_time) %>%
  summarise(mean_wateruptake = mean(wateruptake)) %>%
  ggplot(aes(x = lubridate::as_datetime(date_time),
             y = mean_wateruptake)) +
  geom_point(aes(colour = vegetable_type), size = 4) +
  geom_point(data = enza_long, aes(x = lubridate::as_datetime(date_time), y = wateruptake, colour = vegetable_type), size = 0.2, alpha = 4/5) 

# convert time to numbers
enza_long$date_time_num <- as.numeric(enza_long$date_time) %>%
  abs()

enza_long %>%
  group_by(vegetable_type, date_time_num) %>%
  summarise(mean_wateruptake = mean(wateruptake)) %>%
  ggplot(aes(x = date_time_num,
             y = mean_wateruptake)) +
  geom_point(aes(colour = vegetable_type), size = 4) +
  geom_point(data = enza_long, aes(x =  date_time_num, y = wateruptake, colour = vegetable_type), size = 0.2, alpha = 4/5)

```

## Plotting a logisitc model
```{r}

## linear
glm_model_1 <- glm(log10(wateruptake) ~ poly(date_time_num,1,raw=TRUE) + vegetable_type, data = enza_long)

## second_degree
glm_model_2 <- glm(wateruptake ~ poly(date_time_num,2,raw=TRUE) + vegetable_type, data = enza_long)

## third degree
glm_model_3 <- glm(wateruptake ~ poly(date_time_num,3,raw=TRUE) + vegetable_type, data = enza_long)

## forth degree
glm_model_4 <- glm(wateruptake ~ poly(date_time_num,4,raw=TRUE) + vegetable_type, data = enza_long)


model_list <- list(glm_model_1,
                   glm_model_2,
                   glm_model_3,
                   glm_model_4)

lapply(model_list, summary)

## the AIC does not deacrease from model two to model three, model two is better than model 1 but model 3 and 4 do not increase the fit.

ind <- is.na(enza_long$wateruptake)
missing_date_time <- enza_long[ind, ]

# impute missing data <- missing_date_time$date_time

new_dv <- predict(glm_model_2, newdata = data.frame(date_time_num = missing_date_time$date_time_num, vegetable_type = missing_date_time$vegetable_type))

new_dv

new_data <- tibble(date_time = missing_date_time$date_time, 
                   wateruptake = new_dv, 
                   date_time_num = missing_date_time$date_time_num,                              vegetable_type = missing_date_time$vegetable_type,
                    vegetable = missing_date_time$vegetable)


p <- enza_long %>%
  group_by(vegetable_type, date_time_num) %>%
  summarise(mean_wateruptake = mean(wateruptake)) %>%
  ggplot(aes(x = date_time_num,
             y = mean_wateruptake)) +
  geom_point(aes(colour = vegetable_type), size = 4) +
  geom_point(data = enza_long, aes(x = date_time_num, 
                                   y = wateruptake, 
                                   colour = vegetable_type), 
             size = 0.2, alpha = 4/5)  +
  geom_smooth(data = new_data, (aes(x = date_time_num,
                                   y = wateruptake,
                                   group = vegetable_type)),
             colour = "purple")

p

## Generate full MLM model with interactions


names(new_data)
names(enza_long)


replaced_data <- full_join(na.omit(enza_long), missing_date_time, by = "date_time")
names(replaced_data)
sum(is.na(replaced_data))
missmap(replaced_data, main = "Missing values vs observed")
names(new_data)

p <- new_data %>%
  dplyr::select(date_time, 
                vegetable, 
                vegetable_type,
                wateruptake) %>%
#  group_by(vegetable_type.x, date_time) %>%
#  summarise(mean_wateruptake = mean(wateruptake.x)) %>%
  ggplot(aes(x = date_time,
             y = wateruptake)) +
  geom_point(aes(colour = vegetable_type), 
             size = 0.5,
             shape = 4 ) +
  geom_point(data = enza_long, aes(x = date_time, 
                                   y = wateruptake, 
                                   colour = vegetable_type), 
             size = 0.2, 
             alpha = 3/5,
             shape = 1) 

p



## summary figure 


```

Now we can create a full model with interactions
We fisrt have to set contrasts, in this case only for the vegetable_type variable
```{r}



new_data$vegetable_type %>% as.factor %>%
  levels

prei_tomaat <- c(1, 0, 0)
prei_ui <- c(0, 1, 0)
tomaat_ui <- c(0, 0, 1)

contrasts(new_data$vegetable_type) <- cbind(prei_tomaat,
                                            prei_ui,
                                            tomaat_ui)

contrasts(new_data$vegetable_type)
names(new_data)





veggies_baseline_new <-lme(wateruptake ~ 1, 
                       random = ~1|vegetable/date_time, 
                       data = new_data, method = "ML", 
                       na.action = na.omit)

summary(veggies_baseline_new)

anova(veggies_baseline_new)


## only fixed effects
veggies_fixed_new <- lm(wateruptake ~ vegetable_type, 
                        data = new_data)
AIC(veggies_fixed_new)
AIC(veggies_baseline_new) # select the model with lowest AIC (Akaike Information Criteria)


## add vegetable_type as predictor
veggies_veggie_type_new <-update(veggies_baseline_new, .~. + vegetable_type)

AIC(veggies_veggie_type_new)
anova(veggies_veggie_type_new)

## add date_time as predictor
veggies_type_time_new <-update(veggies_veggie_type_new, .~. + date_time)

AIC(veggies_type_time_new)
anova(veggies_type_time_new)

## add interaction
veggies_full <-update(veggies_type_time_new, .~. + vegetable_type:date_time)

AIC(veggies_full)
anova(veggies_full)


# does not work due to numerical instability of the data:
# Error in MEEM(object, conLin, control$niterEM) : 
#  Singularity in backsolve at level 0, block 1
# try transformations!!
summary(veggies_full)



```

## A final plot of the new data
```{r}

names(replaced_data)
replaced_data %>%
  dplyr::select(date_time,
         wateruptake.x,
         vegetable_type.x,
         vegetable.x) %>%
  group_by(vegetable_type.x, date_time) %>%
  summarise(mean_wateruptake = mean(wateruptake.x)) %>%
  ggplot(aes(x = date_time, y = mean_wateruptake)) +
    geom_point(aes(colour = vegetable_type.x))

```