---
title: "Package `adventr` Statistics tutorials by Andy Field"
author: "Marc A.T. Teunis"
date: '`r Sys.Date()`'
output:
  ioslides_presentation: 
    highlight: pygments
    widescreen: yes
---


```{r, root_1, include=FALSE}
## defines the root of the project for later use
require("rprojroot") || utils::install.packages("rprojroot")
library(rprojroot)
root <- find_root_file(criterion = is_rstudio_project)
```

```{r, knitr_setup, warning=FALSE, error=FALSE, message=FALSE, include=FALSE}
## setting knitr options
if (!require("knitr")) utils::install.packages("knitr")
library(knitr)

knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      error = FALSE, 
                      message = FALSE, fig.width = 5, fig.height = 5)
knitr::opts_knit$set(root.dir = root)
```

## The `learnr` package is a great tool to learn R. 

The author of the book 'Discovering Statitics Using R' (Prof dr. Andy Field) shared his `learnr` tutorials with me. 
In this lesson we will be working with Andy's "Milton Rocks" repository and `learnr` to discover how to properly do ANOVA in R. 

## Packages
```{r, eval=FALSE}
library(tidyverse)
#install.packages("lubridate")
library(devtools)
install_github("profandyfield/adventr")

```

## Contents:

The tutorials are named to correspond (roughly) to the relevant chapter of the book. For example, adventr_03 would be a good tutorial to run alongside teaching related to chapter 3, and so on.

    adventr_02: Data basics in R and RStudio
    adventr_03: Summarizing data (introducing ggplot2)
    adventr_04: Fitting models (central tendency)
    adventr_05: Presenting data (summarizing groups and more ggplot2)
    adventr_08: Inferential statistics and robust estimation (covers Chapter 9 too)
    adventr_11: Hypothesis testing
    adventr_14: The general linear model
    adventr_15: Comparing two means
    adventr_15_rm: Comparing two means (repeated measures)
    adventr_16: Comparing several means
    adventr_16_rm: Comparing several means (repeated measures)
    adventr_17: Factorial designs
    adventr_mlm: Multilevel models (not covered in the book)
    adventr_growth: Growth models (not covered in the book)
    adventr_log: Logistic regression (not covered in the book)

## Running tutorials

To run a tutorial execute:

`learnr::run_tutorial("name_of_tutorial", package = "adventr")`

## Summarizing data
Let's look at "Summarizing data""
```{r, eval=FALSE}
library(adventr)
learnr::run_tutorial("adventr_03", package = "adventr")
```

## Comparing two means
Let's look at "Comparing two means"
```{r, eval=FALSE}
learnr::run_tutorial("adventr_15", package = "adventr")
```

## Multilevel models
```{r, eval=FALSE}
learnr::run_tutorial("adventr_mlm", package = "adventr")
```

## Logistic Regression
```{r, eval=FALSE}
learnr::run_tutorial("adventr_log", package = "adventr")
```
