---
title: "Intro to RStudio and R"
author: "Marc A.T. Teunis"
date: "`r Sys.Date()`"
output:
  ioslides_presentation: 
    highlight: pygments
    widescreen: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      error = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      fig.width = 5, 
                      fig.height = 3)
```

```{r, define_root, include=FALSE}
## defines the root of the project for later use
require("rprojroot") || utils::install.packages("rprojroot")
library(rprojroot)
root <- find_root_file(criterion = is_rstudio_project)
```

```{r, include=FALSE}
library(tidyverse)
```

## Getting to know your neighbour

Spent 4 minutes explaining to your neighbour:

 - What brings you here? What are your expectations?
 - What characterizes you regarding `R`:
 `Beginner`, `Intermediate`, `Advanced`, `Super Man`?
 - With what type of data do you normally work? 
 - What type of problems are you solving, or planning to solve?

## **Discuss with your neighbour:** {.smaller}

**What is wrong with this R code?**

 * Determine  what this code is meant to do
 * Find at leat 3 mistakes
 * What would you do to prevent these kind of mistakes?

```{r, eval=FALSE}
df$var_a[df$var_a ==  999] <- NA
df$var_b[df$var_b == -999] <- NA
df$var_h[df$var_h == -999] <- NA
df$var_i[df$var_j == -989] <- NA
df$var_j[df$var_j == -999] <- NA
df$var_o[df$var_n == -988] <- "NA"
## there are at least 5 errors in this code
```
Reference: Example (adapted from): Hadley Wickham, "Expressing yourself_ with R" Lecture  - https://richmedia.lse.ac.uk/methodologyinstitute/20170814_ExpressYourselfUsingR.mp4

## Functional programming

When working with R you can:

 - Solve each problem seperately (and spend a lot of time debugging and copying pasting)
 - Write small pieces of code (`functions`) that are generalisible to future problems
 
## Writing functions
```{r}
df <- tibble(a = c(1:9, -999), 
             b = c(1:5, -999, -999, 8:10),
             c = LETTERS[1:10])
df
```

## Writing a function that replaces 
`a designated number or string` 
for `NA`

```{r}
replace_x_for_na <- function(x){
  
  x[x == -999] <- NA
  return(x)
}
```

## Apply this function to one column:
```{r}
replace_x_for_na(df$a)
```

## Apply our function to our whole df
```{r}
df <- purrr::modify_if(df, is.numeric, replace_x_for_na)
df
```

## Good functions:

 * Do one thing good
 * Have no 'side-effects'
 * `class(input) = class(output)` (to make the function 'pipe-pable')
 
__You will hear about the 'pipe (`%>%`)' later__ 

__This workshop does not cover `writing functions` in dept__

We hope this small example got you inspired in writing more of your own functions

# Let's create some R objects
### Switch to RStudio RMarkdown document "part1_intro_to_rstudio_and_r.Rmd"

## _'numeric'_ vectors
```{r}
numbers <- c(1.1:10.5)
numbers
sum(numbers)
integer_vector <- c(1:5)
integer_vector
```

## What happens to a 'mixed' vector?
```{r, mixed}
vector <- c(1:10, NA, NA, NA,  3, "Python is great!", "R is greater!")
vector
missing <- is.na(vector) %>% ## which is NA
  print() ## prints to the Console
```

## Type of object
```{r}
class(numbers)
class(missing) ## what is the `class`?
class(vector)
class(integer_vector)
```

## Getting amount of `NA`s
```{r}
missing_num <- as.numeric(missing) ## turn logical into numeric  
missing_num
sum(missing_num) ## how many TRUEs
sum(missing) ## implicitly turning  logical into numeric
sum(is.na(vector)) ##  how many NAs
```

## Mathematical operations on a `character` vector
```{r, eval=FALSE}
sum(vector) ## Warning sum on character
```

## Let's create some data 
```{r}
set.seed(1234)
normals <- rnorm(n = 100, mean = 5, sd = 1.3)
df <- normals %>% as_tibble()
## change number of observations to 10000 and rerun
```

## Plot histogram
```{r}
df %>% ## pipe `%>%` takes a value an puts it in a function
  ggplot(aes(x = value)) +
  geom_histogram(bins = 50)
## change the `bins` argument to see what happens
```

## **Discuss with your neighbour:**

 * How would you turn the above plot in a function that takes two arguments: `bins` and `df`?
 * Try it out using this backbone:
 
 ```
plot_histogram <- function(df, bins){

 ## write plot code here, use the example above
 
 plot <- df %>% ggplot(....)
 ...
 ...
 ...

 return(plot)

}
 
```


## R is case sensitive

*_Important note: R is case-sensitive, meaning that it will matter whether you use capital or lower case, keep this in mind if you get a cryptic warning or error!_*

## Programming (in R) can be frustrating {.build}

 * Writing (R) code can be challenging 
 * You will get mind-boggling errors. 
 * R is extremely picky about capitals and every opening `(`, `[` or `{` must be matched by a closing `)`, `]` or `}`. 

Mind that in R `.` is the default decimal placeholder!  

## Getting help {.build}

 * If you're stuck, try the help (`?` or `??` or `help()`) function. 
 * You can get help about any R function by running `?function_name` and `?package_name` 
 * or `??package_name`, skip down to the examples and look for code that matches what you're trying to do.
 * Another great tool is `Google` or `Stack Overflow`: trying googling the error message, as it's likely someone else has had the same problem, and has gotten help online.

## Statistics and R {.build}

 * This workshop does not cover statistical applications of R
 * All be it, we will look at an example of a linear regression later on
 * Nowadays R is a versatile language and can be used for various applications, but it was originally designed as a language for performing statistical analyses 
 * This is why very many statistical applications are available for R
 * To start learning about statisics and R, I can highly recommend the book "Discovering Statistics Using R" by Dr. Andy Field:
 
https://uk.sagepub.com/en-gb/eur/discovering-statistics-using-r/book236067%20

## For more on learning R, start e.g. with:

http://www.statsteachr.org/ or

https://www.coursera.org/learn/statistical-inference/home/welcome or

https://www.youtube.com/watch?v=ACWuV16tdhY&index=21&list=PLqzoL9-eJTNBDdKgJgJzaQcY6OXmsXAHU and

https://www.youtube.com/watch?v=kvmSAXhX9Hs&index=29&list=PLqzoL9-eJTNBDdKgJgJzaQcY6OXmsXAHU 

and the rest of Mike Marin's lectures, that are also a really great way to start learning R.
Or try datacamp.com

# Packages

## List of available CRAN packages 
http://cran.r-project.org/web/packages/available_packages_by_date.html

For this workshop we mainly use the `{tidyverse}` 
```{r, echo=FALSE}

knitr::include_graphics(path = file.path(root,
                                         "images",
                                         "tidyverse_sticker.png"))
```
```{r}
# install.packages("tidyverse")
library(tidyverse)
```

## `{tidyverse}` {.build}
See: http://tidyverse.tidyverse.org/ for more info
The `{tidyverse}` package includes (among others):

 * Wrangling: `{dplyr}`, `{tidyr}` 
 * Reading data: `{readr}`, `{readxl}`
 * Functional prgramming: `{purrr}`
 * Models: `{broom}`
 * Strings and regular expressions: `{stringr}`
 * Factors: `{forcats}`
 * Time (series): `{lubridate}`

For all full list type `tidyverse_packages()` in the console, try it now!

## 'Biological' Packages
For 'Biological' application go to http://www.bioconductor.org

Installing Bioconductor packages is different from CRAN packages
example: `{affy}` 

Each Bioconductor package has a landing page:

https://www.bioconductor.org/packages/release/bioc/html/affy.html

```{r, eval=FALSE}
## installing `{affy}` package
BiocManager::install("affy")
biocLite("affy")

## loading affy package and the vignettes pages of this package
# library(affy)
# browseVignettes("affy")
```

## Vignettes {.build}

 * Vignettes are long explanations and demos of a package
 * A vignette contains examples and a workflow that shows how the package can be used and which (research) questions can be adressed with the functions and datasets in the package
 * It usually is a good place to start for examples
 * It also shows the so-called dependencies of a package
 * It explains which other packages you would need and how the data should look to be able to work with the package the vignette belongs to

## Getting Help examples
```{r, eval=FALSE}
#install.packages("dplyr")
library(dplyr)
library(ggplot2)
?dplyr
?ggplot2
?mean
?mean  # goes to the page with functions related to '.mean.'
apropos("mean") # search on more options of or alternatives for a certain function
formals(dplyr::filter)
```

## Examples and demos on functions and packages
```{r, eval=FALSE}
example(mean) # to see a worked example
demo(graphics) # demonstration of R functions
```

## Functions; naming arguments
```{r, error=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
seed <- c(1:4)
set.seed(seed = seed)
q <- rnorm(n = 10000, mean = 20, sd = 2)
hist(q)
```

## No naming
```{r, eval=FALSE}
set.seed(seed)
qq <- rnorm(10000, 20, 2)
all(q == qq)
```

__**Write code for humans AND computers, be explicit!**__

# R (Data) objects

## Data Structures: Lists

When using R for data analysis you will most likely work with data in a matrix, an array, a list or even more likely: a dataframe.

A matrix is a table with only numeric values. An array consists of multiple matices. A list is collection of R objects of different data type. A dataframe is a table with variable names in the first row and observations in the consecutive rows. The columns in a dataframe represnet different variables.  

The dataframe and the list are the most widely used datastructures when considering experimental data. 

## Create a List
```{r}
lst <-list(first_names = c(male = "Fred", female = "Mary"), no.children = 3, child_ages=c(4,7,16),
child_names = c("Suzy", "Marvin", "Jane"), address = c("Pandamonium Alley 114, Chaosville"),
marital_status = TRUE)
length(lst) # number of elements in the list
lst
```

## Accessing items in a list
```{r}
lst[[1]] # 1st element of List
lst[[3]][2] # second item of third element
names(lst) # named elements in this list
lst$child_names # pull "named" elements from a list using `$` operator
```

## `str()` gives you the structure of an object
```{r}
str(lst) # display structure of lst
``` 
The new RStudio Interface also enables interactive exploration of R-objects (demo)

## Selecting single elements in a list
To select a single element from a variable in a list
```{r}
lst$child_ages[3] 
lst[[6]][2] # returns the value of the second element for your variable
``` 

## Looping over lists
```{r}
purrr::map(lst, is.na)
```

## The Dataframe
The dataframe is the most widely used data structure 
**Remember "Tidy data!"**   

## Creating a data frame from individual vectors
```{r}
people_df <- data.frame(age = c(24, 27, 19, 34),      
                        sex = c("F","F","M", "M"), 
                        weight = c(64, 55, 80, 70),
                        names = c("Christa", "Suzan", 
                                  "Matt", "John"))
head(people_df)
``` 

## Or the 'tidyverse' way
```{r}
people_tbl <- tibble::tibble(age = c(24, 27, 19, 34),      
                             sex = c("F","F","M", "M"), 
                             weight = c(64, 55, 80, 70),
                             names = c("Christa", "Suzan", 
                                       "Matt", "John"))
people_tbl
```

## Viewing the contents of a dataframe
```{r, eval=FALSE}
summary(people_df)
table(people_df)

head(people_df) 			# gives the content of the data frame
names(people_df) 
str(people_df)

people_df$age # gives the content of the variable "age" from the data frame

``` 

## Using index on dataframes
Using the index "[]" on a dataframe is a bit tricky. The dataframe always consists of rows and columns. Indexing a dataframe goes like:

`dataframe[row number(s), column number(s)]`

```{r, eval=FALSE}
people_df$age[1] 	# first element of this vector
people_df[,2] 	  # content of 2nd variable (column) which is a character                      vector -> factor
people_df[1,] 	  # content of the 1st row
                  # multiple indices
people_df[2:3, c(1,3)] # remember to use c

```

## Import data into R
```{r}
library(tidyverse)
path_to_gender_age_data <- file.path(root, "data", "gender.txt")
gender_age <- read_delim(path_to_gender_age_data,
                         delim = "/")
```

## Inspecting the dataframe
```{r}
# getting the first few rows 
head(gender_age)
# getting information on the variable, the dimensions
str(gender_age)
# selecting a variable
gender_age$gender
gender_age$age
gender_age[1:6,2]
``` 

## read_csv
CSV is a format of a data file that uses commas or semicolons as separators for the columns.

```{r}
library(readr)
skin <- read_csv(paste0(root, "/data/skincolumns.csv")) 
skin %>% head(3)
```

## Inspecting the skin dataframe
```{r, eval=FALSE}
head(skin)	 # content of the data frame
dim(skin)
attributes(skin)
summary(skin)
## ?read_csv 	 # help on the function
```

## `skin` dataset contains an NA, some functions do not work with NAs:
```{r}
mean(skin$`Genotype A`)
mean(skin$`Genotype B`)

# to remove the NA, take care: consider leaving NAs in and use arguments like na.rm = TRUE
skin_noNA <- na.omit(skin)
mean(skin_noNA$`Genotype B`)
``` 

## Let's clean up the workspace
```{r, eval = FALSE }
rm(list=ls())
root <- find_root_file(criterion = is_rstudio_project)
## Note: never use this in code that is meant for others!!!
```

*_The above is an effective way to clear all the items in the Global Environment, but is is not very friendly to use this in code you share with others: can you think of why?_*

## Vectors {.build}
R is an object oriented language: meaning you can create and work with (manipulate/index/access) objects. Vectors are R's elementary objects and 
come in different flavours:

 - Nummeric vector: contains only numbers: decimal separator in R is "." (decimal point) and not "," (decimal comma) as is common in the English language. 
 - Character vector: contains only "words", but words can also be numbers: "23" or other items "100%" or "$2,000.00" 
 - An integer vector: an nummeric series: 1, 2, 3 is an integer of length 3.
 - Logical: logical vectors contain only two values: "TRUE" and/or "FALSE"
 - Mixed: Vectors do not need to be of one type. They can be mixed. They can only be of one class, so this operation will induce **_coercion_**.

## Numeric vectors
```{r}
c(2,8,5) 	# combines its arguments to form a vector
nv_1 <- c(2,8,5) # assignment statement
nv_1 # view the content of the object "nv_1"
```

## Class tells you about the type of the variable
```{r}
nv_2 <- c(8.4,5.6,10.1,13.1,2.5,7.8,15.2,3.8,20.9)
nv_2
class(nv_2)
```

## **Discuss with your neighbour**

 * What happens to the vector values if you run ```nv_2_int <- as.integer(nv_2)``` 
 * Call ```class(nv_2_int)``` and ```nv_2_int``` to confirm your suspecion
 * What is the apparent difference between "nummeric" and "integer" vectors? 

# Number notations and rounding

## Scientific notations 
```{r}
big_numbers <- rnorm(10, mean = 10000000, sd = 2000000)
big_numbers %>% formatC(format = "e", digits = 5)

many_digits <- c(2.55858868688584848)
round(many_digits, digits = 3)
sqrt(many_digits * 1000 /200 * 2^6)  

```

## Rounding numbers
```{r}
small_numbers <- runif(10, min = 0.001, max = 0.1) %>% print()
small_numbers %>% round(digits = 2)
```

## Character vectors
```{r, eval=FALSE}
cv <- c("this is", "an", "example of", "1", "character", "vector", "with length:", "length(cv)")
cv
sum(cv)

length(cv)
class(cv)
```

## Logical vectors
```{r}
lv <- c(TRUE, FALSE, TRUE, TRUE)
lv
class(lv)
## logical vectors can also be converted to numeric vectors
nlv <- as.numeric(lv)
nlv
## note that coercion of a logical to a numeric vector changes the "TRUE" value to 1 and the "FALSE" value to 0
```

## Manipulating vectors
You can add/subtract/devide or use other arithmetic functions on numeric vectors
```{r} 
a <- c(1,3,5,7,9)
b <- c(2,4,6,8)

z1 <- a - b
z1

z2 <- b - a
z2
z3 <- a / b
z3
```

## Apply functions to a vector 
```{r}
z4 <- sum(a)
z4
z5 <- max(a) - max(b)
z5
```   

## Make sequences
```{r}
repetitions <- rep("Luke", times = 10)
saber <- rep("Sword", length.out = 20)
jedi <- c(repetitions, saber) %>% rep(each = 3)
```

## Indexing with [...]
You can get the individual items of a vector by using the index `[]`
```{r}
x <- c(8, 5, 10, 13, 2, 7, 15, 3, 20, 8)
x # create vector with 10 variables
length(x)
mode(x) # information on data mode (numeric, character, logic)

```

## Using the index
```{r}
x
x[3] ## creating a subset by indexing:
x[c(3, 4, 7)] # apply a simple function

mean(x[c(2,5)]) # example of a function
```

## Using the vector index `[]` some more
```{r}
i <- 5
x[c(i, i+2)] # (i = 5 & i = 7), i-th element
x[-2] # all but the ith (second) element
x[3:5] # element 3 to 5
x[x > 9] # all greater than some value
```

## Concatenate or `c()` command
The `c()` function can be used to __concatenate__, __combine__ or __create__ objects in R 
```{r}
pp <- c(x, 0, x)
length(pp) 
sum(c(pp, c(1:5), c(1,2, c(1:10)))) # sum of elements in the 'concatenated' vector
```

## Generate R-code from this Rmd
```{r, eval=FALSE}
knitr::purl("hands_ons/part1_intro_to_rstudio_and_r.Rmd", output = "./hands_ons/part1_intro_to_rstudio_and_r.R")
```
