---
title: "First EncounteR"
author: "Marc A.T. Teunis"
date: '`r Sys.Date()`'
output: 
   BiocStyle::html_document:
    toc: true
    toc_depth: 2
---
<style>

.footer {
    font-size: 16pt;
    color: Black;
    font-family: 'Helvetica'
    content: attr(label);
    position: absolute;
    margin-left: auto;
    margin-right: auto;
    left: 0;
    right: 0;
    text-align:center;
    bottom: 10px;
    line-height: 1.9;
    display: block;
}

</style>

<div class="footer" style="margin-top;font-size:120%;"> 
Using R in Life Sciences - January 2018 </div>

```{r, root_1, include=FALSE}
## defines the root of the project for later use
require("rprojroot") || utils::install.packages("rprojroot")
library(rprojroot)
root <- find_root_file(criterion = is_rstudio_project)
```

```{r, knitr_setup, warning=FALSE, error=FALSE, message=FALSE, include=FALSE}
## setting knitr options
if (!require("knitr")) utils::install.packages("knitr")
library(knitr)

knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      error = FALSE, 
                      message = FALSE, fig.width = 5, fig.height = 5)
knitr::opts_knit$set(root.dir = root)
```

## Packages
```{r}
library(tidyverse)
```

# Intro to R and RStudio "On the fly Demo"


## The data 

 - The datafile "demo_populations.csv" contains over two million rows:
First try to import the file "./data/demo_populations.csv"
into Excel, what happens?
 
 - Try making a summary box plot for each country, for the two sampled populations
in Excel

## In MS Excel
```{r, echo=FALSE, out.width="1000px"}
knitr::include_graphics(path = file.path(root, 
                                         "images", 
                                         "importing_big_file_in_excel.jpg"))
```

## Importing the same file in R
```{r}
populations <- read_delim(file = file.path(root, "data", 
                                           "demo_populations.csv"), 
                                            delim = ",")
head(populations, 4)
## number of rows:
formatC(nrow(populations), format = "e", digits = 1)
```

## Graphs - R code
```{r, dpi=130}
## get names of variables
names(populations)

## summarize data
bars <- populations %>%
  group_by(countries, sample) %>%
  summarise(mean_value = mean(value)) %>% 
  # the above result is piped (using %>%) to the ggplot() function

## graph data using ggplot2
ggplot(aes(x = sample, y = mean_value)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ countries)
```

## Bar Graphs - images
```{r}
bars
```

## R as a calculator
```{r calculator}
1 + 1 # add
sqrt(2) # root square 
2^2 # raise to the power of
(900 - 20 + (90*20)/2)^60 # more complex with ()
```

## R does statistics
```{r random numbers}
set.seed(123)
```

Let's assume we have a numeric measure for two population samples (1.3 million) 
in five countries each. 
So each country was sampled twice (the data we used above). 
Performing a t.test is easy
```{r generate_samples}
pop_1 <- rnorm(mean = 4.89, sd = 0.89, n = 1300000)
pop_2 <- rnorm(mean = 10.34, sd = 2.83, n = 1300000)
t.test(pop_1, pop_2)
```

Is the T-test the right test here?

## There are many packages to add-on to R
```{r packages, eval = FALSE}
library(tidyverse)
library(RforProteomics)
library(readMzXmlData)
library(pheatmap)
library(DESeq2)
library(stringr)
library(drc)

```

## R likes dataframes
```{r coerce_to_dataframe}
populations <- tibble(pop_1, pop_2)
head(populations, 4)
```

## You can add variables (the 5 countries)
```{r add_countries_variable}
countries <- rep(1:5, times = length(pop_1)/5)
populations <- tibble(pop_1, pop_2, countries) %>%
  print()

```

To see the dataframe
```{r head_dataframe}
populations
```

## R can reshape data
```{r reshape}
populations <- populations %>% 
  gather(key = sample, value = value, pop_1:pop_2) %>%
  print()
```

## R can plot
```{r plot_freqpoly}
ggplot(data = populations, aes(x = value)) +
  geom_freqpoly(aes(color = sample))
```


## R can summarize data
```{r data_summary}
populations$sample <- as.factor((populations$sample))
levels(populations$sample)

summary_populations <- populations %>% 
  group_by(sample) %>%
  summarise(mean = mean(value),
            sd = sd(value))
```


## A more meaningful bar plot
```{r plot_bar, fig.width=4, fig.height=4}
ggplot(data = summary_populations, aes(x = sample, y = mean)) + 
  geom_bar(aes(fill = sample), stat = "identity", position = "dodge") +
  geom_errorbar(aes(ymin = mean-sd, ymax = mean+sd),
                width=.2,                    # Width of the error bars
                position=position_dodge(.9))
```


## Save graphics
```{r ggsave, eval=FALSE}
ggsave(filename = "./images/demo_bar.png")
```

## Save dataframe
```{r save_data, eval=FALSE}
write_csv(populations, path = "./data/demo_populations.csv")
```

## R can read external data
```{r read_data}
gene_expression <- read_csv(file.path(root, "data", "results.csv"))
gene_expression
```

## There are many resources
```{r resources, eval = FALSE}
# ?dplyr
# ?filter
# ?dplyr::filter
url_blog <- "https://www.r-bloggers.com/"
url_doc <- "https://www.rdocumentation.org/"
url_cran <-"https://www.r-project.org/help.html" 
url_hadley <- "http://hadley.nz/"
```

## You can create you own functions
```{r functions}
?rep

## function:
  repeating <- function(string, times){
  rep <- rep(string, times = times)
  rep <- as_tibble(rep)
  names(rep) <- c("input_string")
  return(rep)
    
 }
```

## Apply `repeating()` on string
```{r}
r_is_great <- c("R is great for Data Science")
## apply string to function
r_is_great_table <- repeating(r_is_great, times = 100)
r_is_great_table
```

## Looping over lists with functions
Let's apply the function above on multiple strings in a list
```{r}

## First we create a list with strings

strings_list <- as.list(c("These are the first step to greatness with R",
                          "Functional programming in R is pretty difficult!",
                          "Once you get to know R, your life will never be the same",
                          "R is way better than Matlab!!"))

## inspect the list
strings_list

```

## Apply the function `repeating()` to our `strings_list`
```{r}

repeating_many <- lapply(strings_list, repeating, times = 10) %>%
##  convert to tibble
dplyr::bind_rows() %>%
  as_tibble() %>%
  group_by(input_string) %>%
  summarise(count = n())

repeating_many
## look at the ordering, does it make sense?
```
