---
title: "Bioconductor Workflows"
author: "Marc A.T. Teunis"
date: "`r Sys.Date()`"
output:
  ioslides_presentation: 
    highlight: pygments
    widescreen: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5, fig.height = 3, eval = FALSE)
```

```{r, root_1, include=FALSE}
## defines the root of the project for later use
require("rprojroot") || utils::install.packages("rprojroot")
library(rprojroot)
root <- find_root_file(criterion = is_rstudio_project)
```

## Demos and Workflows
BIOCONDUCTOR offers a number of workflows to illustrate packages in relation to analysis of Biological data. There are many different workflows available 

Examples in this course:

 - Biological Sequences
 - RNA sequencing
 - Oligonucleotide Arrays  
 - Gene annotation
 - Proteomics / LC-MS data

## Getting workflows:
To view the workflows go to https://www.bioconductor.org/help/workflows/ 

## Sequence data
An exemplary workflow that introduces a number of BIOCONDUCTOR R packages to anlyze sequence data.
https://www.bioconductor.org/help/workflows/sequencing/ 

```{r, eval=FALSE}
BiocManager::install("sequencing")
workflowInstall("sequencing")
```

## Vignettes
Vignettes are manuals and explanations how packages and workflow can be used. They usually demonstrate packages or workflows by example or case from the literature.

Some workflows have vignettes. Try `browseVignettes("Workflow_Name")`

To see vignettes of packages try:
```{r}
help(package="GenomicRanges")
vignette(package="GenomicRanges")
```

## RNA sequencing
To install the workflows for RNAsequencing:
```{r}
BiocManager::install("RNAseq123")
```

## Oligonucleotide Arrays  
the workflow for oligonucleotide array data analysis:
```{r}
BiocManager::install("arrays")
```

## Gene annotation
For gene annotation several workflow are available:
```{r}
BiocManager::install("annotation")
```

and 
http://bioconductor.org/help/workflows/annotation/Annotating_Genomic_Ranges/ 
that uses the above workflow.

and a second workflow on Annotating Genomic Variants 
```{r}
BiocManager::install("variants")
```

## Proteomics / LC-MS data
For proteomics we also have several workflows available.
```{r}
## external dependencies 
## sudo apt-get install libnetcdf-dev

install.packages("ncdf4")
BiocManager::install("RforProteomics")
browseVignettes("RforProteomics")
library(RforProteomics)

```

and 
```{r}
BiocManager::install("proteomics")
```

## Exercises

1) Install the workflow for "arrays"

