---
title: "R basics - Exercise"
author: "Marc A.T. Teunis"
date: "`r Sys.Date()`"
output: 
   BiocStyle::html_document:
    toc: true
    toc_depth: 2
---

# Exercise
To complete this exercise you will need to upload a Rmd and corresponding rendered HTML file containing:

 - The answers to the exercises
 
 Upload the Rmd and rendered HTML files in CANVAS at the appropriate exercise number. 


# This document is an R Markdown Document
An example of the use of _literate programming_ in R. R Markdown is based on the Markdown mark-up language used for easy writing for the web and can combine plain human-readable text with computer code and output from that code in one document. 

If you want to hide the code you can set the option in the _setup_ block to `ECHO = FALSE`.

# Knitr options
Knitr is an R package that translates Rmd and LaTeX files into human-readable documents. In the first lesson we already saw the document that contains the code chunk like below. This chunk takes care of setting some options for the knitr package: We will explain later what this code does.
```{r setup}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      error = FALSE,
                      results = 'hide',
                      fig.show = 'hide')
``` 

# Packages
```{r, results='markup', echo=TRUE}
library(tidyverse)
```

# Setting the project root
We need to an object for the root of the project for purposes yet to explain.
```{r, echo=TRUE, eval=TRUE, results='markup'}
if (!require("rprojroot")) install.packages("rprojroot")
library("rprojroot")
root <- find_root_file(criterion = is_rstudio_project)
root
```

# Load example dataset
```{r, echo=TRUE, eval=TRUE, results='markup'}
library(readr)
simple_IO <- read_csv(paste0(root, "/data/simple-IO.csv"))
dim(simple_IO)
ncol(simple_IO)
nrow(simple_IO)
```

1) Inspecting a dataframe

 A) How do we inspect `simple_IO`?
 B) How many variables does `simple_IO` have?
 C) How many rows?
 D) call `summary()` on `simple_IO`; In which variable do we observe the highest expression ratio?

2) Pasting objects

# To illustrate the difference between functions paste0() and paste()
```{r, echo=TRUE}
a <- c("a", "b", "c", NA)
b <- c(1:4)
c <- c(6:8, NA)
d <- c(1.3, 1.6, 6.7)
e <- as.integer(d)
f <- as.integer(a)
g <- as.numeric(a)

(z <- paste(a, b, c))
(zz <- paste0(a, b, c))

```

 A) What does the above `as.integer()` function do? And `as.numeric()`?

 B) Check the contents of `e`, `f` and `g`.

# Exercises R structures & objects

3) R as a calculator

R can be used as a "regular" calculator, we will practice this a bit, just to get familiar with the way R works and displays its information

3A) What is the sum of 3, 12 and 250?
```{r}
sum(3, 12, 250)
```

3B) What is the difference between 4902 and 3987?
```{r}

4902 - 3987

```

3C) Calculate 2 to the power of 24
```{r}

2^24

```

3D) What is 240 divided by 14?
```{r}

240/14

```

3F) What is the average of 12 and 20?
```{r}
mean(c(12, 20))
## or
(12 + 20)/2

```
3G) What is the average of 10.8, 13.7 and 15.9?

```{r}

mean(c(10.8, 13.7, 15.9))

```

# Data types & structures
R has a variety of different data types and structures. Knowing how to work with them is fundamental to learning and using R. The following exercises give a basic introduction to creating and manipulating them.
Using the `c()` function here that `creates` vectors is essential.

4A) Create a vector consisting of the numbers 10.5, 20.4, 30.5, 28.9 and 5,4.
```{r}

v <- c(10.5, 20.4, 30.5, 28.9, 5.4)

```

4B)	Create a vector consisting of a decreasing sequence of integers from 50 to -10.
```{r}

v <- c(50:-10)

```

4C) Create a vector consisting of an increasing sequence of even integers from -10 to 50. You can use the `seq()` function here.
```{r}

v <- seq(-10, 50, 2)

```

4D) Add the number 5 to the tail of the vector in 4C, and save the result as a new vector.
```{r}

v2 <- c(v, 5)

```
	
4E) Create a vector of five hundred alternating zeros and ones. Use the function `rep()` here.
```{r}

v <- rep(c(0, 1), 500)

```

4F) Create a vector of a hundred, uniformly distributed, random values between -1 and +1. (Hint: see the function `runif()` )
Use `set.seed(123)` to set the random number generator.
```{r}
?rnorm

set.seed(123)
v <- runif(100, min= -1, max= 1)
hist(v)

```

4G)	Calculate the average of these random values. (Hint: see the function `mean()`. 

Which value do you think the mean will be approaching. Discuss with you neighbour before running the code!!
```{r}

mean(v)

```

4H)	Now replace position 5 in the vector with NA, use `v[5]` in you call. 
```{r}
v[5] <- NA 
## Have a look at the result
v
```

4I)	Calculate the mean again. (Hint: make sure you deal with the missing values!). Look at `?mean()` for the solution.
```{r}

mean(v, na.rm=TRUE)

```


# The Matrix
5A) Create a matrix of four columns and three rows with increasing integers from 1 to 12, left to right and then top to bottom. Use the function `matrix()`. See the help first.
```{r}

m <- matrix(1:12, nrow=3, byrow=TRUE)

```

5B)	Create a matrix of four columns and three rows, the first column only containing 1, the second only 2, the third only 3 and the fourth column only 4.
```{r}

m <- matrix(rep(c(1, 2, 3, 4), 3), nrow=3, byrow=TRUE)
m
```

5C)	Calculate the sum of this matrix. (Hint: see the function sum() )
```{r}

sum(m)

```

# Vectors

6A) Create the following items:

 * A numerical vector named "nr" consisting of the following numbers: 1.8, 4.5, 10.1, 8.3 and 7.5
```{r}

nr <- c(1.8, 4.5, 10.1, 8.3, 7.5)

``` 

 * A numerical vector named "prime" consisting of the following integers: 1, 3, 5, 7 and 11
```{r}

prime <- c(1, 3, 5, 7, 11)

```

 * A character vector named "name" consisting of the following values: abc1, foo2, bar3 and app5
```{r}

name <- c("abc1", "foo2", "bar3", "app5")

```

 * A logical vector named "valid" consisting of: TRUE, FALSE, FALSE, TRUE and FALSE
```{r}
valid <- c(TRUE, FALSE, FALSE, TRUE, FALSE)
valid
```

# The dataframe (tibble)

7A) Try to create a `tibble` named "mydata" with the function `tibble()` from the above four vectors. This will fail, why? .
```{r, eval=FALSE}
# Not all vectors are of equal length.
mydata <- tibble(nr, prime, name, valid)

```

7B) Fix the error you found and create the dataframe. Also try giving the columns different names. (Hint: see the function `names()` and the function `tibble()`.
```{r}
# Not all vectors are of equal length; addingan addiotnal element

name <- c("abc1", "foo2", "bar3", "app5", NA)

mydata <- tibble(nr, prime, name, valid)
names(mydata) <- c("nr", "prime", "name", "valid")

```

7C) What happens if you try to calculate the sum of the data.frame?
```{r, eval=FALSE}

# You get an error, something like "only defined 
# on data frame with all #numerical variables"
sum(mydata)

```

7D) When you look at the structure of the tibble using `str()`, you will see that the column with the "abc1", "foo2", etc values has a certain data type. Which data type is "name"? Do you think it is the right type? Or do you need to change it?

```{r}
str(mydata)
#The variable "name" is actually a factor, change the type of this variable if necessary.

mydata$name <- as.factor(mydata$name)
mydata
```

Note: the new packages of the tidyverse (readr and readxl) deal with this problem. Check out the documentation here:
https://github.com/tidyverse/tidyverse 

