---
title: "Exercise - Data Wrangling"
author: "Marc A.T. Teunis"
date: "`r Sys.Date()`"
output: 
   BiocStyle::html_document:
    toc: true
    toc_depth: 2
---

```{r setup}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      error = FALSE,
                      results = 'hide',
                      fig.show = 'hide')
``` 

# End product

**Generate a new RMarkdown file containing the answers to all questions below and render it to HTML. Upload your _Rmd AND HTML_ file to CANVAS under the corresponding (see file name) exercise number** 

# Prerequisites

 - Go over all Dplyr demo's of UNIT 2 (4 tutorials in total)
 - Review the chapter "Relational Data" in "R for Data Science" by Grolemund and Wickham 

# Packages
```{r, results='markup', echo=TRUE}
library(tidyverse)
library(jsonlite)
library(tidyverse)
```

# Setting the project root
We need an object for the root of the project for robust and reproducible code.
```{r, echo=TRUE, eval=TRUE, results='markup'}
if (!require("rprojroot")) install.packages("rprojroot")
library("rprojroot")
root <- find_root_file(criterion = is_rstudio_project)
root
```

# Prequisites:

 - For this exercise you need to review the demos for `{dplyr}`
 - Review the `join()` functions here: https://r4ds.had.co.nz/relational-data.html 
 - Use dplyr syntax as much as possible
 - You need to review visualizations with `{ggplot2}` to be able to create the requested graphs: https://r4ds.had.co.nz/data-visualisation.html
 - Use the `%>%` operator where you can
 - Prevent creating new variables in the `.GlobalEnv` as much as possible
 
# Data
This data was obtained from: https://www.kaggle.com/nationalparkservice
The dataset consists of two files in a zip archive:

# Question 1. Getting the data

The files are located in `"./data/biodiversity-in-national-parks.zip"`
and are called:
 
 - `parks.csv`
 - `species.csv`

A) Extract the data using the `unzip()` function (this only need to be perfomed once)

```{r, echo=FALSE}
unzip(zipfile = file.path(root, "data", "biodiversity-in-national-parks.zip"),
      exdir = file.path("data"))

library(here)


readr::read_csv(here("data", "airway_counts.csv"))


```

B) Read both data files into R

Name them `species_data` and `park_data` appropriately according to the orignal file names.
```{r}
species_data <- read_csv(file = file.path(root,
                                           "data",
                                           "species.csv"))

park_data <- read_csv(file = file.path(root,
                                           "data",
                                           "parks.csv"))

```

# Question 2 - `filter()`

A) How many different species (`Scientific Name`) are there in `species_data`?
```{r}
## dplyr
names(species_data)
species_data %>% 
   select(`Scientific Name`) %>%
   unique() %>%
#   arrange() %>%
   count()

## base
species_data$`Scientific Name` %>%
   as_factor() %>%
   levels() %>%
   length()
```

B) How many parks are there in `park_data`?
```{r}
names(park_data)
## dplyr
park_data %>%
   select(`Park Name`)%>%
   unique() %>%
   count()
```

C) Using the filter function, how many species have the label "Endangered"
```{r}
species_data %>%
   dplyr::filter(`Conservation Status` == "Endangered") %>%
   select(`Scientific Name`) %>%
   unique() %>%
   count()
```

D) Which park has the most endangered species?

**TIPS**

 - You can use the `group_by()` function here, together with `tally(sort = TRUE)` 
 
```{r}
species_data %>%
   dplyr::filter(`Conservation Status` == "Endangered") %>%
   group_by(`Park Name`) %>% 
   select(`Scientific Name`, `Park Name`) %>%
   tally(sort = TRUE)
    
```

# Qestions 3 - Join & Clean

A) Join the `species_data` and the `parks_data` together by using a `left_join()`. You need to retain all the species data. Put the joined data in a new variable in the `.GlobalEnv`, name it `joined_data`.

```{r}
joined_data <- left_join(species_data, park_data)
```

**Use the `joined_data` table for the rest of the exercise**

B) Set the variable names to lower case and rplace the spaces in the name for a "_" using `str_replace_all()`
```{r}
names(joined_data) <- names(joined_data) %>% 
   tolower() %>%
   str_replace_all(string = ., pattern = " ", replacement = "_")
head(joined_data,3)
```

## Qestion 4 - Mutate
A) The surface area of the Parks are provided in Acres. Create a new variable called `surface_km2` that displays the surface area in km^2 (square km). Try to provide scientific notation.
see for more details: https://stackoverflow.com/questions/39623636/forcing-r-output-to-be-scientific-notation-with-at-most-two-decimals/39625148

the conversion for acres to km^2 is:

$km^2 = ac / 247.11$

**TIPS**
Do not forget to store the new column in the data

```{r}
joined_data <- joined_data %>%
   mutate(surface_km2 = acres/247.11,
          surface_km2 = formatC(surface_km2, format = "e", digits = 2))

joined_data$surface_km2 %>% max()
```

# Question 5 - Summarize
**Create a new table called `summary_data`**

A) Calculate the total number of parks that **each** species is reported in. Create a new table and a new variable inside that table (`parks_recorded`). Sort from high to low.
```{r}
summary_data <- joined_data %>%
   group_by(scientific_name) %>%
   summarise(parks_recorded = n()) %>%
   arrange(parks_recorded)

summary_data
```

B) Create a new variable in the `summary_table` that displays the total surface area for each species in this table. Call this variable `total_surface`. Use the `surface_km2` variable to get to the answer.

**TIPS**

```{r}

summary_data <- joined_data %>%
   group_by(scientific_name) %>%
   mutate(surface_km2 = as.numeric(surface_km2)) %>%
   summarise(number_species = n(),
             surface_area = sum(surface_km2)) %>%
   arrange(surface_area)
summary_data
```


C) Which species (multiple) have the smallest area to live in?

```{r}

summary_data
map(summary_data, class) == "numeric"

summary_data %>%
   dplyr::filter(surface_area == 22.5) %>%
   arrange(surface_area) %>%
   select(scientific_name) %>%
   unique()
```

# Question 6 - Visualize
A) Create a plot that visualizes the relationship between the surface area of the parks `surface_km2` and the number of "Endangered" or "Species of Concern" animals

```{r}
x <- joined_data %>%
   dplyr::filter(conservation_status == "Endangered" |
                 conservation_status == "Species of Concern") %>%
   mutate(surface_km2 = as.numeric(surface_km2)) %>%
   group_by(park_name, conservation_status) %>%
   summarize(n = n(),
             surface = mean(surface_km2)) 
x

names(park_data) <- names(park_data) %>% tolower() %>%
   str_replace_all(string = ., pattern = " ", replacement = "_")

   
left_join(x, park_data)   


#   summarize(number_problem_species = n(),
#             total_surface = sum(surface_km2)) 

x %>%
   ggplot(aes(x = surface,
              y = n)) +
   geom_point(aes(colour = conservation_status))
```

B) Try to solve the overplotting 
```{r}
x %>%
   ggplot(aes(x = log10(surface),
              y = n)) +
   geom_point(aes(colour = conservation_status)) 
```

C) Add a linear model
```{r}
x %>%
   ggplot(aes(x = log10(surface),
              y = n)) +
   geom_point(aes(colour = conservation_status)) +
   geom_smooth(aes(group = conservation_status), method = "lm")
```

```{r}
# install.packages("odbc")
library(odbc)
library(dplyr)
library(dbplyr)
library(DBI)
library(purrr)
suppressWarnings(library(rlang, warn.conflicts = FALSE)) 
```


# Question 7. Creating or Connecting to a database

Follow the workflow in the link below and answer the questions below:

https://bioconductor.org/packages/release/bioc/vignettes/Organism.dplyr/inst/doc/Organism.dplyr.html

```{r}
# BiocManager::install("Organism.dplyr")
```

A) Find all the transcripts for the genes that carry `FOXP` (Forkhead receptor) in their name (`Symbol`) in B. taurus?

**TIPS**
You need to install the appropriate `TxDb` organism package, before you can you use `Organism.dplyr` for that species. 
```{r}

library(TxDb.Btaurus.UCSC.bosTau8.refGene)
library(Organism.dplyr)

# BiocManager::install("TxDb.Btaurus.UCSC.bosTau8.refGene")
# BiocManager::install("org.Bt.eg.db")

## Running src_organism() without a given path will save the sqlite file to a tempdir():
src <- src_organism("TxDb.Btaurus.UCSC.bosTau8.refGene")

#Alternatively you can provide explicit path to where the sqlite file should be saved, and re-use the data base at a later date.
#path <- "path/to/my.sqlite"
#src <- src_organism("TxDb.Hsapiens.UCSC.hg38.knownGene", path)

#supportedOrganisms() provides a list of organisms with corresponding ‘org’ and ‘TxDb’ packages being supported.

supportedOrganisms()

src_tbls(src)

tbl(src, "ranges_gene")

tbl(src, "id") %>%
    dplyr::filter(symbol %like% "FOXP%") %>%
    dplyr::select(entrez, alias, ensembl, symbol) %>%
    distinct() %>% arrange(symbol) %>% collect()


go <- inner_join(tbl(src, "id"), tbl(src, "id_go")) %>%
    filter(symbol %like% "FOXP%") %>%
    dplyr::select(entrez, ensembl, symbol, go, evidence, ontology) %>%
    collect(n=Inf)

go
txcount <- inner_join(tbl(src, "id"), tbl(src, "ranges_tx")) %>%
    filter(symbol %like% "FOXP%") %>%
    dplyr::select(symbol, tx_id, ensembl, entrez) %>%
    group_by(symbol) %>%
 #   summarize(count = n()) %>%
 #   dplyr::select(symbol, count) %>%
 #   arrange(desc(count)) %>%
    collect(n=Inf)

txcount
ids <- txcount$entrez
ids

```


B) What are the `uniprot` ids belonging to the list of transcripts (`entrez`) ids of A)
```{r, eval=FALSE}
### needs work

src_tbls(src)

tbl(src, "id_protein")


proteins <- inner_join(tbl(src, "id_transcript"), tbl(src, "id_protein")) %>%
    filter(entrez %in% ids) %>%
    dplyr::select(uniprot, ensembltrans, entrez) %>%
#    group_by(symbol) %>%
#   summarize(count = n()) %>%
 #   dplyr::select(symbol, count) %>%
 #   arrange(desc(count)) %>%
    collect(n=Inf)
proteins
```


https://mdsr-book.github.io/instructor/sql.html

